---
title: "Registered patients compared with ONS population estimates"
subtitle: "Focused national and LSOA-level analysis"
output:
  word_document: default
  html_document:
    toc: true
    toc_float: true
params:
  start_date: "2015-07-01"
  end_date: "2026-08-31"
  snapshot_date: "2024-07-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  dpi = 150
)

required_packages <- c(
  "DBI", "odbc", "dplyr", "tidyr", "ggplot2", "scales",
  "knitr", "httr", "jsonlite", "readr", "glue", "rlang", "tibble",
  "tidyverse"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the following R packages before rendering: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(DBI)
library(odbc)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(knitr)

start_date <- as.Date(params$start_date)
end_date <- as.Date(params$end_date)
snapshot_date <- as.Date(params$snapshot_date)

stopifnot(
  !is.na(start_date),
  !is.na(end_date),
  !is.na(snapshot_date),
  start_date <= end_date
)

pct_difference <- function(registered, ons) {
  ifelse(
    is.na(ons) | ons == 0,
    NA_real_,
    100 * (registered - ons) / ons
  )
}

# Build the pound sign at render time to avoid escaped Unicode text in charts.
gbp_symbol <- intToUtf8(163)

comparison_bar <- function(
  data,
  category,
  title,
  subtitle,
  y_title = NULL,
  fixed_fill = NULL
) {
  category_quo <- rlang::enquo(category)

  p <- ggplot(
    data,
    aes(
      x = Pct_Difference,
      y = !!category_quo
    )
  ) +
    geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.5)

  if (is.null(fixed_fill)) {
    p <- p +
      geom_col(aes(fill = Pct_Difference), width = 0.7) +
      scale_fill_gradient2(
        low = "#B2182B",
        mid = "#F2F2F2",
        high = "#2166AC",
        midpoint = 0,
        guide = "none"
      )
  } else {
    p <- p + geom_col(fill = fixed_fill, width = 0.7)
  }

  p +
    geom_text(
      aes(label = sprintf("%.1f%%", Pct_Difference)),
      hjust = ifelse(data$Pct_Difference >= 0, -0.12, 1.12),
      size = 3.7
    ) +
    scale_x_continuous(
      labels = function(x) paste0(x, "%"),
      expand = expansion(mult = c(0.16, 0.18))
    ) +
    coord_cartesian(clip = "off") +
    labs(
      title = title,
      subtitle = subtitle,
      x = "Percentage difference from ONS",
      y = y_title,
      caption = "Percentage difference = (registered patients − ONS estimate) / ONS estimate × 100."
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = margin(5.5, 30, 5.5, 5.5)
    )
}


# HTML tables start closed; document outputs retain ordinary visible tables.
report_table <- function(x, caption = NULL, ...) {
  if (!knitr::is_html_output()) {
    return(knitr::kable(x, caption = caption, ...))
  }
  summary <- if (is.null(caption)) "View table" else paste("View table:", caption)
  summary <- gsub("&", "&amp;", summary, fixed = TRUE)
  summary <- gsub("<", "&lt;", summary, fixed = TRUE)
  summary <- gsub(">", "&gt;", summary, fixed = TRUE)
  table_html <- paste(
    knitr::kable(x, format = "html", caption = caption, ...),
    collapse = "\n"
  )
  knitr::asis_output(paste0(
    "\n<details class=\"report-table\"><summary>", summary, "</summary>\n",
    "<div class=\"report-table-scroll\">\n", table_html,
    "\n</div>\n</details>\n"
  ))
}

first_matching_column <- function(data, exact = character(), regex = NULL) {
  nm <- names(data)

  exact_match <- exact[exact %in% nm]
  if (length(exact_match) > 0) {
    return(exact_match[[1]])
  }

  if (!is.null(regex)) {
    regex_match <- nm[grepl(regex, nm, ignore.case = TRUE)]
    if (length(regex_match) > 0) {
      return(regex_match[[1]])
    }
  }

  NA_character_
}
```


```{r report-table-style, echo=FALSE, results='asis'}
if (knitr::is_html_output()) {
  cat('<style>
.report-table { margin: 1em 0; border: 1px solid #d8dfe5; border-radius: 4px; padding: .6em .8em; }
.report-table > summary { cursor: pointer; font-weight: 600; color: #2166ac; }
.report-table > summary:focus-visible { outline: 2px solid #2166ac; outline-offset: 3px; }
.report-table-scroll { overflow-x: auto; margin-top: .8em; }
.report-table-scroll table { width: 100%; }
</style>')
}
```

```{r connect-and-load-data, include=FALSE}
source("01_connect_to_facts_and_dimensions.R")
source("03_Payments_To_GP_Practice.R")
source("08_Get_Focused_Patient_List_Data.R")
source("09_Get_LSOA_NHS_Region_Lookup.R")
source("10_Get_Practice_Region_Lookup.R")
DBI::dbDisconnect(con)
source("12_Summarise_Registration_Coverage.R")
```

```{r prepare-analysis-data, include=FALSE}
age_levels <- c(
  "0-9", "10-19", "20-29", "30-39", "40-49",
  "50-59", "60-69", "70-79", "80-89", "90+"
)

national_time_series <- bind_rows(national_registered, national_ons)

demographic_time_series <- full_join(
  practice_registered_demographic_time_series,
  ons_demographic_time_series,
  by = c("Period", "Sex", "Age_Band")
) %>%
  mutate(
    Age_Band = factor(Age_Band, levels = age_levels),
    Sex = factor(Sex, levels = c("Female", "Male")),
    Pct_Difference = pct_difference(Registered, ONS)
  )

sum_if_available <- function(x) {
  if (all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
}

age_time_series <- demographic_time_series %>%
  group_by(Period, Age_Band) %>%
  summarise(
    Registered = sum_if_available(Registered),
    ONS = sum_if_available(ONS),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Registered, ONS),
    names_to = "Source",
    values_to = "Population"
  ) %>%
  mutate(
    Source = recode(
      Source,
      Registered = "Registered patients",
      ONS = "ONS population estimate"
    )
  )

sex_time_series <- demographic_time_series %>%
  group_by(Period, Sex) %>%
  summarise(
    Registered = sum_if_available(Registered),
    ONS = sum_if_available(ONS),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Registered, ONS),
    names_to = "Source",
    values_to = "Population"
  ) %>%
  mutate(
    Source = recode(
      Source,
      Registered = "Registered patients",
      ONS = "ONS population estimate"
    )
  )

national_snapshot <- national_time_series %>%
  filter(Period == snapshot_date) %>%
  select(Source, Population) %>%
  pivot_wider(names_from = Source, values_from = Population) %>%
  transmute(
    Geography = "England",
    Registered = `Registered patients`,
    ONS = `ONS population estimate`,
    Pct_Difference = pct_difference(Registered, ONS)
  )

age_sex_snapshot <- demographic_time_series %>%
  filter(Period == snapshot_date)

age_snapshot <- age_sex_snapshot %>%
  group_by(Age_Band) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Pct_Difference = pct_difference(Registered, ONS))

sex_snapshot <- age_sex_snapshot %>%
  group_by(Sex) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Pct_Difference = pct_difference(Registered, ONS))

lsoa_july_all <- full_join(
  lsoa_registered_july,
  lsoa_ons_july,
  by = c("LSOA_Code", "Period")
) %>%
  mutate(
    Registered = replace_na(Registered, 0),
    ONS = replace_na(ONS, 0),
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

# Explicit England-resident subset for every LSOA-derived analysis.
lsoa_july_england <- lsoa_july_all %>%
  filter(grepl("^E01", LSOA_Code))

# Quantify registered patients excluded from the England-resident LSOA subset.
non_england_registered_snapshot <- lsoa_registered_july %>%
  filter(
    Period == snapshot_date,
    !grepl("^E01", LSOA_Code)
  ) %>%
  mutate(
    Residence_Group = case_when(
      grepl("^W0", LSOA_Code) ~ "Wales",
      TRUE ~ "Other non-English LSOA"
    )
  ) %>%
  group_by(Residence_Group) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    .groups = "drop"
  )

registered_removed_wales <- non_england_registered_snapshot %>%
  filter(Residence_Group == "Wales") %>%
  summarise(Registered = sum(Registered, na.rm = TRUE)) %>%
  pull(Registered)

registered_removed_other <- non_england_registered_snapshot %>%
  filter(Residence_Group == "Other non-English LSOA") %>%
  summarise(Registered = sum(Registered, na.rm = TRUE)) %>%
  pull(Registered)

registered_removed_non_england <-
  registered_removed_wales + registered_removed_other

lsoa_registered_snapshot_total <- lsoa_registered_july %>%
  filter(Period == snapshot_date) %>%
  summarise(Registered = sum(Registered, na.rm = TRUE)) %>%
  pull(Registered)

registered_removed_non_england_pct <-
  100 * registered_removed_non_england / lsoa_registered_snapshot_total

if (nrow(anti_join(
  lsoa_july_england, nhser_lookup, by = c("LSOA_Code" = "LSOA21CD")
)) > 0) {
  stop("Some English LSOAs are missing from the April 2026 geography lookup.")
}

lsoa_july_geography <- lsoa_july_england %>%
  inner_join(nhser_lookup, by = c("LSOA_Code" = "LSOA21CD"))

registration_coverage <- summarise_registration_coverage(
  national_registered, lsoa_registered_july, imd_lookup, snapshot_date
)

imd_july <- lsoa_july_england %>%
  inner_join(
    imd_lookup %>% mutate(
      IMD_Decile = suppressWarnings(as.numeric(as.character(IMD_Decile)))
    ),
    by = "LSOA_Code"
  ) %>%
  filter(IMD_Decile %in% 1:10) %>%
  group_by(Period, IMD_Decile) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS),
    IMD_Label = factor(
      IMD_Decile,
      levels = 10:1,
      labels = c(
        "10 (Least deprived)", "9", "8", "7", "6",
        "5", "4", "3", "2", "1 (Most deprived)"
      )
    )
  )

region_july <- lsoa_july_geography %>%
  group_by(Period, NHSER26CD, NHSER26NM) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

icb_july <- lsoa_july_geography %>%
  group_by(Period, ICB26CD, ICB26NM) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

imd_snapshot <- imd_july %>% filter(Period == snapshot_date)
if (abs(sum(imd_snapshot$Registered) - registration_coverage$imd_patients) > 0.5) {
  stop("The deprivation analysis does not reconcile with the opening coverage summary.")
}
region_snapshot <- region_july %>% filter(Period == snapshot_date)
icb_snapshot <- icb_july %>% filter(Period == snapshot_date)

required_practice_lookup_cols <- c("Practice_Code", "NHS_Region")
missing_region_lookup_cols <- setdiff(
  required_practice_lookup_cols,
  names(practice_region_lookup)
)

if (length(missing_region_lookup_cols) > 0) {
  stop(
    "practice_region_lookup is missing required columns: ",
    paste(missing_region_lookup_cols, collapse = ", ")
  )
}

required_icb_lookup_cols <- c("Practice_Code", "ICB_Key", "ICB")
missing_icb_lookup_cols <- setdiff(
  required_icb_lookup_cols,
  names(practice_icb_lookup)
)

if (length(missing_icb_lookup_cols) > 0) {
  stop(
    "practice_icb_lookup is missing required columns: ",
    paste(missing_icb_lookup_cols, collapse = ", ")
  )
}

if (nrow(practice_icb_lookup) == 0) {
  stop(
    "The practice-to-ICB lookup is empty. Check the practice ICB fields ",
    "in payments2425.csv and 10_Get_Practice_Region_Lookup.R."
  )
}
```

# Registration sources and deprivation coverage

```{r opening-registration-summary-prepare, include=FALSE}
coverage_difference_text <- function(value, reference, reference_label) {
  difference <- value - reference
  if (difference == 0) {
    return(paste0("the same number of registrations as the ", reference_label, " total"))
  }
  paste0(
    scales::comma(abs(difference), accuracy = 1),
    if (difference < 0) " fewer registrations than the " else " more registrations than the ",
    reference_label, " total (",
    sprintf("%.2f%%", 100 * abs(difference) / reference),
    " of that reference total)"
  )
}
```

For `r format(snapshot_date, "%d %B %Y")`, the practice-level source contains **`r scales::comma(registration_coverage$practice_total)` registered patients**, compared with **`r scales::comma(registration_coverage$lsoa_total)`** in the LSOA-level source before any residence filter or deprivation linkage. The LSOA source therefore contains `r coverage_difference_text(registration_coverage$lsoa_total, registration_coverage$practice_total, "practice-level")`. This source difference arises before any restriction to English LSOAs or linkage to IMD. [NHS England reported processing issues in the July 2024 LSOA release](https://digital.nhs.uk/data-and-information/publications/statistical/patients-registered-at-a-gp-practice/july-2024) and warned that its totals might differ from practice totals.

The LSOA extract includes **`r scales::comma(registration_coverage$welsh_patients)` registered patients across `r scales::comma(registration_coverage$welsh_lsoa_count)` Welsh LSOAs** (`r sprintf("%.2f%%", registration_coverage$welsh_patients_pct)` of all LSOA registrations). Welsh LSOAs do not have values in the English Index of Multiple Deprivation (IMD) used here: [Wales has a separate Welsh Index of Multiple Deprivation, whose deciles are not directly comparable with England's](https://www.gov.uk/government/statistics/english-indices-of-deprivation-2025/english-indices-of-deprivation-2025-frequently-asked-questions). Those Welsh registrations are therefore excluded from the English deprivation analysis. Another `r scales::comma(registration_coverage$other_patients)` registrations have other or unassigned residence codes and are also excluded, leaving **`r scales::comma(registration_coverage$english_patients)` registrations in English LSOAs**. Welsh registrations are already excluded by the England-residence filter and are counted once in this reduction.

Within the English subset, **`r scales::comma(registration_coverage$english_patients_without_imd)` registrations in `r scales::comma(registration_coverage$english_lsoas_without_imd)` LSOAs** have no matching valid IMD decile and are excluded at the deprivation-linkage stage. `r if (registration_coverage$english_patients_without_imd == 0) "There is no further reduction in registered-patient counts from IMD linkage within the English subset." else paste0("This is an additional reduction of ", sprintf("%.2f%%", registration_coverage$english_imd_excluded_pct), " of English-LSOA registrations, beyond the residence exclusions.")`

Consequently, the deprivation analyses contain **`r scales::comma(registration_coverage$imd_patients)` registered patients**: `r coverage_difference_text(registration_coverage$imd_patients, registration_coverage$lsoa_total, "unfiltered LSOA-level")`, and `r coverage_difference_text(registration_coverage$imd_patients, registration_coverage$practice_total, "practice-level")`. The reduction from the unfiltered LSOA total is the sum of Welsh-residence exclusions, other/unassigned-residence exclusions and the additional English IMD-linkage exclusions. The comparison with the practice total also includes the initial practice-versus-LSOA source difference. These are differences in aggregate registration counts, not evidence identifying individual ghost patients.

## Checking residence codes and IMD linkage

The other/unassigned-residence count is calculated from the registration source before joining IMD. Changing the IMD edition cannot assign a deprivation decile to a registration with no usable residence LSOA. Welsh residence, other/unassigned residence and English LSOAs failing IMD linkage are separate exclusions. The tables below show both the reasons and the actual unmatched codes so these counts can be checked without combining them.

```{r registration-imd-source-date, results='asis'}
imd_source_dates <- sort(unique(as.Date(imd_lookup$IMD_Source_Date)))
imd_source_date_text <- if (length(imd_source_dates) == 0L || anyNA(imd_source_dates)) {
  "unavailable"
} else {
  paste(format(imd_source_dates, "%d %B %Y"), collapse = ", ")
}
cat(paste0("The IMD lookup uses the latest warehouse snapshot: **",
           imd_source_date_text, "**. This date alone does not identify the IMD edition or LSOA geography.\n"))
```

[IMD 2019 uses 2011 LSOAs; IMD 2025 uses 2021 LSOAs](https://assets.publishing.service.gov.uk/media/68ff59c80f801e57b5bef907/ID_2025_Technical_Report.pdf). The July 2024 registration source uses 2021 LSOAs. If retaining the deprivation measure available in 2024, use IMD 2019 with a documented 2021-to-2011 geography mapping; [ONS has used such a mapping to assign 2019 deprivation classifications to Census 2021 LSOAs](https://www.ons.gov.uk/peoplepopulationandcommunity/healthandsocialcare/healthandwellbeing/articles/generalhealthbyagesexanddeprivationenglandandwales/census2021). A crosswalk changes the geography, not the IMD edition. Splits or mergers need a documented allocation and must not duplicate registration counts or average decile labels.

Alternatively, a retrospective analysis can join 2024 registrations directly to IMD 2025 on matching 2021 LSOA codes, but must label the result as **2024 registrations classified by IMD 2025**, rather than deprivation measured in 2024. The two index editions are not interchangeable. This report does not automatically convert between them or fill missing residence with guessed deciles. Remaining valid English codes absent from the lookup should be investigated for geography, formatting or source-coverage problems.

```{r registration-imd-linkage-summary, results='asis'}
registration_coverage$linkage_audit %>%
  group_by(Linkage_Status) %>%
  summarise(
    `Residence-code rows` = n(),
    `Registered patients` = sum(Registered),
    .groups = "drop"
  ) %>%
  mutate(`Registered patients` = scales::comma(`Registered patients`)) %>%
  rename(`Linkage status` = Linkage_Status) %>%
  report_table(caption = "Residence exclusions and English IMD linkage status")
```

```{r registration-imd-unmatched-codes, results='asis'}
registration_coverage$linkage_audit %>%
  filter(!Linkage_Status %in% c(
    "English LSOA: valid IMD decile",
    "Welsh residence: separate deprivation index"
  )) %>%
  arrange(desc(Registered), LSOA_Code) %>%
  transmute(
    `LSOA/residence code` = if_else(is.na(LSOA_Code), "<missing>", LSOA_Code),
    `Registered patients` = scales::comma(Registered),
    `Reason excluded` = Linkage_Status,
    `Lookup IMD decile` = if_else(is.na(IMD_Decile), "Unavailable", as.character(IMD_Decile))
  ) %>%
  report_table(caption = "Other/unassigned residence and English codes without valid IMD")
```

# Scope and interpretation

This report compares GP registered-patient counts with ONS population estimates for England from July 2015 onwards. ONS estimates are restricted to England. National and age/sex registered-patient series are practice-based and therefore include all patients registered with English GP practices. LSOA-derived analyses explicitly restrict registered patients to English residence LSOAs before IMD, NHS region and ICB aggregation. Percentage difference is defined as:

\[
100 \times \frac{\text{registered patients} - \text{ONS population}}{\text{ONS population}}
\]

Positive values mean registered-patient counts exceed ONS estimates. Snapshot analyses use `r format(snapshot_date, "%B %Y")`.

In the HTML report, select **View table** to expand a table beneath the charts. Tables remain visible in the Word report.

# National population over time

```{r national-population-time-series}
ggplot(
  national_time_series,
  aes(x = Period, y = Population, colour = Source)
) +
  geom_line(linewidth = 1.1) +
  geom_point(
    data = national_time_series %>%
      filter(Source == "ONS population estimate"),
    size = 2
  ) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(
    limits = c(start_date, end_date),
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "m")
  ) +
  labs(
    title = "England population: registered patients and ONS estimates",
    subtitle = "July 2015 onwards; available source snapshots",
    x = NULL,
    y = "Population",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Age over time

Age bands contain completed ages 0–9, 10–19, and so on, with 90+ as the final band. Both sources use the same boundaries. The ONS curves include English residence LSOAs only; Welsh residents are excluded. Registered counts include patients of English GP practices, irrespective of residence.

```{r age-time-series, fig.height=10}
ggplot(
  age_time_series,
  aes(x = Period, y = Population, colour = Source, group = Source)
) +
  geom_vline(
    xintercept = as.numeric(as.Date("2021-07-01")),
    linetype = "dashed", colour = "grey65", linewidth = 0.4
  ) +
  geom_line(
    data = age_time_series %>% filter(!is.na(Population)),
    linewidth = 0.8
  ) +
  geom_point(
    data = age_time_series %>%
      filter(Source == "ONS population estimate", !is.na(Population)),
    size = 1.3
  ) +
  facet_wrap(~ Age_Band, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "m")) +
  labs(
    title = "Population by age over time",
    subtitle = "English GP registrations and England-only ONS estimates; matching ten-year age bands",
    x = NULL,
    y = "Population",
    colour = NULL,
    caption = "Dashed line: Census 2021 year; the cause of any turning point is not established by the chart."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

## Checking the mid-2024 age totals

The [published ONS mid-2024 workbook](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/estimatesofthepopulationforenglandandwales/mid20242023localauthorityboundaries) gives **6,795,519** people aged 0–9 in England and Wales: **6,473,967 in England** and **321,552 in Wales**. The England figure is the relevant geographic comparator for this report. Including age ten would instead give 7,169,868 for England, so a 0–10 total must never be labelled or compared as 0–9.

The expandable table compares the database's July 2024 snapshot with the published mid-2024 reference for every band. A database snapshot date does not by itself establish the ONS reference year or publication vintage. If the England totals differ after using the same age boundaries, check the warehouse's estimate year, revisions, completeness and duplicate records. Do not replace database values simply to force agreement with the benchmark. The national age/sex extracts are also checked against their own source totals so excluded age or sex records cannot disappear silently.

```{r ons-mid2024-age-reference, results='asis'}
ons_mid2024_reference <- readr::read_csv(
  "data/ons_mid2024_age_band_reference.csv", show_col_types = FALSE
)
ons_database_2024 <- ons_demographic_time_series %>%
  filter(Period == as.Date("2024-07-01")) %>%
  group_by(Age_Band) %>%
  summarise(Database_England = sum(ONS), .groups = "drop")

if (nrow(ons_database_2024) > 0L) {
  ons_mid2024_reference %>%
    left_join(ons_database_2024, by = "Age_Band") %>%
    transmute(
      `Age band` = Age_Band,
      `Database England, July 2024` = scales::comma(Database_England),
      `Published England, mid-2024` = scales::comma(England),
      `Published Wales, mid-2024` = scales::comma(Wales),
      `Published England and Wales` = scales::comma(England_and_Wales),
      `Database minus published England` = scales::comma(Database_England - England)
    ) %>%
    report_table(caption = "Database age totals versus the published ONS mid-2024 reference")
} else {
  cat("No July 2024 demographic snapshot is available in the selected database extract.\n")
}
```

## Interpreting the apparent turn around 2021

The change in the 0–9 and 20–29 curves needs to be considered alongside both demography and revisions to the population estimates. [ONS mid-2021 estimates introduced the Census 2021 population base](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/bulletins/annualmidyearpopulationestimates/mid2021), and [ONS subsequently rebased the pre-2021 series](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/bulletins/rebasingofmidyearpopulationestimatesfollowing/latest). Combining older pre-census estimates with newer post-census estimates can create a discontinuity. The database extract does not identify the publication vintage of each year's estimate, so a census-related break is a possibility to check, rather than a confirmed explanation for these particular curves.

For **ages 20–29**, continued growth after 2021 is consistent with increased net international migration and changes in the cohorts entering and leaving the age band. [ONS identified migration as the main contributor to population growth in the year to mid-2023 and noted large increases among 20–29-year-olds in many rapidly growing local authorities](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/bulletins/populationestimatesforenglandandwales/mid2023). This supports migration as a plausible contributor, but does not quantify how much of this report's national age-band change it explains.

For **ages 0–9**, the population is a stock of ten birth cohorts, changed by births, children ageing out at ten, deaths and migration of children. A rise therefore does not necessarily mean that births have started increasing. [Births in England and Wales increased briefly in 2021](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/bulletins/birthsummarytablesenglandandwales/2021), then [fell in 2022](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/bulletins/birthsummarytablesenglandandwales/2022) and [2023](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/bulletins/birthsummarytablesenglandandwales/2023). ONS rebasing also reduced, rather than universally increased, estimates for children in England. The apparent upturn should therefore be checked against a consistently revised ONS series and child migration/cohort flows before it is attributed to a sustained change in births.

The expandable audit below reports the actual July estimates and changes for the two age bands, together with the number of English LSOAs represented in each source snapshot. Changes in LSOA counts can reflect new boundaries or incomplete coverage; a boundary change alone should not change the national total if coverage is complete. Annual changes are shown only where consecutive years are present. These checks identify the timing of the observed change; they do not establish its cause.

```{r ons-age-trend-audit, results='asis'}
ons_age_trend_audit <- age_time_series %>%
  filter(
    Source == "ONS population estimate",
    Age_Band %in% c("0-9", "20-29"),
    format(Period, "%m") == "07",
    !is.na(Population)
  ) %>%
  mutate(Year = as.integer(format(Period, "%Y"))) %>%
  arrange(Age_Band, Period) %>%
  group_by(Age_Band) %>%
  mutate(
    Previous_Population = lag(Population),
    Consecutive_Year = Year == lag(Year) + 1L,
    Annual_Change = if_else(Consecutive_Year, Population - Previous_Population, NA_real_),
    Annual_Change_Pct = if_else(
      Consecutive_Year & Previous_Population > 0,
      100 * Annual_Change / Previous_Population, NA_real_
    )
  ) %>%
  ungroup() %>%
  left_join(national_ons %>% select(Period, LSOA_Count), by = "Period") %>%
  filter(Year >= 2018) %>%
  transmute(
    Year, `Age band` = Age_Band,
    `ONS population` = scales::comma(Population),
    `Annual change, n` = if_else(is.na(Annual_Change), "Not available", scales::comma(Annual_Change)),
    `Annual change, %` = if_else(is.na(Annual_Change_Pct), "Not available",
                                 sprintf("%.2f%%", Annual_Change_Pct)),
    `English LSOAs in source, n` = scales::comma(LSOA_Count)
  )
report_table(ons_age_trend_audit, caption = "ONS age-band estimates and source coverage around 2021")
```

# Sex over time

```{r sex-time-series, fig.height=7}
ggplot(
  sex_time_series,
  aes(x = Period, y = Population, colour = Source, group = Source)
) +
  geom_line(
    data = sex_time_series %>% filter(!is.na(Population)),
    linewidth = 1
  ) +
  geom_point(
    data = sex_time_series %>%
      filter(Source == "ONS population estimate", !is.na(Population)),
    size = 1.8
  ) +
  facet_wrap(~ Sex, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "m")) +
  labs(
    title = "Population by sex over time",
    subtitle = "Registered patients and ONS estimates",
    x = NULL,
    y = "Population",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

# July 2024 national snapshots

## National total

```{r national-snapshot-chart}
comparison_bar(
  national_snapshot,
  Geography,
  "National percentage difference",
  format(snapshot_date, "%B %Y"),
  fixed_fill = "#2166AC"
)
```

```{r national-snapshot-table, results='asis'}
national_snapshot %>%
  transmute(
    Geography,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "National population comparison")
```

## GP practice totals compared with GP LSOA totals

This compares both registered-patient sources at the same snapshot date. The GP LSOA total includes all residence codes, including Wales and unassigned/other records, before any ONS join or England-only filter. A positive difference means the practice-based total is larger. The difference alone does not identify ghost patients.

```{r practice-versus-lsoa-table, results='asis'}
practice_total_snapshot <- national_registered %>%
  filter(Period == snapshot_date)

if (nrow(practice_total_snapshot) != 1L ||
    nrow(lsoa_registered_july %>% filter(Period == snapshot_date)) == 0L) {
  stop("Both GP sources must have data for the requested snapshot date.")
}

practice_lsoa_comparison <- practice_total_snapshot %>%
  transmute(
    Period,
    Practice_Total = Population,
    LSOA_Total = lsoa_registered_snapshot_total,
    Difference = Practice_Total - LSOA_Total,
    Pct_Difference = if_else(
      Practice_Total > 0, 100 * Difference / Practice_Total, NA_real_
    )
  )

practice_lsoa_comparison %>%
  transmute(
    Snapshot = format(Period, "%d %b %Y"),
    `GP practice total` = scales::comma(Practice_Total),
    `GP LSOA total (all records)` = scales::comma(LSOA_Total),
    `Practice minus LSOA` = scales::comma(Difference),
    `Difference (% of practice total)` = sprintf("%.2f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Overall difference between GP practice and GP LSOA list totals")
```

Of the GP LSOA total, `r scales::comma(lsoa_registered_snapshot_total - registered_removed_non_england)` patients have an English LSOA, `r scales::comma(registered_removed_wales)` have a Welsh LSOA, and `r scales::comma(registered_removed_other)` have another or unassigned residence code. Later LSOA analyses use the English-LSOA subtotal.

## Age

```{r age-snapshot-chart, fig.height=6}
comparison_bar(
  age_snapshot,
  Age_Band,
  "Percentage difference by age",
  format(snapshot_date, "%B %Y"),
  "Age band"
)
```

```{r age-snapshot-table, results='asis'}
age_snapshot %>%
  transmute(
    `Age band` = Age_Band,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by age")
```

## Sex

```{r sex-snapshot-chart}
comparison_bar(
  sex_snapshot,
  Sex,
  "Percentage difference by sex",
  format(snapshot_date, "%B %Y"),
  "Sex"
)
```

```{r sex-snapshot-table, results='asis'}
sex_snapshot %>%
  transmute(
    Sex,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by sex")
```

## Age and sex

```{r age-sex-snapshot-chart, fig.height=7}
ggplot(
  age_sex_snapshot,
  aes(x = Age_Band, y = Pct_Difference, fill = Sex)
) +
  geom_hline(yintercept = 0, colour = "grey55") +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(
    aes(
      label = sprintf("%.1f%%", Pct_Difference),
      group = Sex,
      vjust = if_else(Pct_Difference >= 0, -0.35, 1.35)
    ),
    position = position_dodge(width = 0.8),
    size = 3.4
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0.12, 0.14))
  ) +
  coord_cartesian(clip = "off") +
  scale_fill_manual(values = c(Female = "#CC79A7", Male = "#0072B2")) +
  labs(
    title = "Percentage difference by age and sex",
    subtitle = format(snapshot_date, "%B %Y"),
    x = "Age band",
    y = "Percentage difference from ONS",
    fill = "Sex"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r age-sex-snapshot-table, results='asis'}
age_sex_snapshot %>%
  arrange(Age_Band, Sex) %>%
  transmute(
    `Age band` = Age_Band,
    Sex,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by age and sex")
```

# IMD deciles using LSOA data


```{r imd-snapshot-chart, fig.height=6}
comparison_bar(
  imd_snapshot,
  IMD_Label,
  "Percentage difference by LSOA-level IMD decile",
  format(snapshot_date, "%B %Y"),
  "IMD decile"
)
```

**England-resident LSOA scope:** all LSOA-derived analyses from this point use only English `E01...` LSOAs. At the `r format(snapshot_date, "%B %Y")` snapshot, this excludes **`r scales::comma(registered_removed_non_england)` registered patients** living in non-English LSOAs: **`r scales::comma(registered_removed_wales)` in Wales** and **`r scales::comma(registered_removed_other)` in other non-English LSOAs** (`r sprintf("%.2f%%", registered_removed_non_england_pct)` of registered patients in the LSOA dataset).

```{r imd-snapshot-table, results='asis'}
imd_snapshot %>%
  arrange(IMD_Decile) %>%
  transmute(
    `IMD decile` = IMD_Label,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by IMD decile")
```

# NHS England regions

```{r region-snapshot-chart, fig.height=6}
region_snapshot_plot <- region_snapshot %>%
  mutate(NHSER26NM = reorder(NHSER26NM, Pct_Difference))

comparison_bar(
  region_snapshot_plot,
  NHSER26NM,
  "Percentage difference by NHS England region",
  format(snapshot_date, "%B %Y"),
  "NHS England region"
)
```

```{r region-snapshot-table, results='asis'}
region_snapshot %>%
  arrange(desc(Pct_Difference)) %>%
  transmute(
    `NHS England region` = NHSER26NM,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by NHS England region")
```

# NHS integrated care boards

All ICB comparisons use the April 2026 structure. The population snapshot remains July 2024. LSOA populations are reassigned through the official ONS April 2026 lookup, including the Frimley boundary changes. Practice counts, workforce and payments use the same 2026 ICB codes.

```{r icb-snapshot-chart, fig.height=12}
icb_snapshot_plot <- icb_snapshot %>%
  mutate(ICB26NM = reorder(ICB26NM, Pct_Difference))

comparison_bar(
  icb_snapshot_plot,
  ICB26NM,
  "Percentage difference by NHS integrated care board",
  format(snapshot_date, "%B %Y"),
  "Integrated care board"
)
```

```{r icb-snapshot-table, results='asis'}
icb_snapshot %>%
  arrange(desc(Pct_Difference)) %>%
  transmute(
    ICB = ICB26NM,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by NHS integrated care board")
```


# Patients per qualified GP FTE

```{r patients-per-gp-prepare, include=FALSE}
workforce_national_july <- qualified_gp_workforce %>%
  filter(format(Period, "%m") == "07") %>%
  mutate(Year = as.integer(format(Period, "%Y"))) %>%
  group_by(Year) %>%
  summarise(GP_FTE = sum(GP_FTE, na.rm = TRUE), .groups = "drop")

population_national_july <- national_time_series %>%
  filter(format(Period, "%m") == "07") %>%
  mutate(Year = as.integer(format(Period, "%Y"))) %>%
  select(Year, Source, Population)

patients_per_gp_july <- population_national_july %>%
  inner_join(workforce_national_july, by = "Year") %>%
  mutate(
    Period = as.Date(paste0(Year, "-07-01")),
    Patients_Per_GP_FTE = Population / GP_FTE
  )

workforce_snapshot <- qualified_gp_workforce %>%
  filter(format(Period, "%Y-%m") == format(snapshot_date, "%Y-%m"))

workforce_region_snapshot <- workforce_snapshot %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(GP_FTE = sum(GP_FTE, na.rm = TRUE), .groups = "drop")

practice_region_registered <- practice_registered_snapshot %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(Registered = sum(Registered, na.rm = TRUE), .groups = "drop")

patients_per_gp_region_snapshot <- region_snapshot %>%
  transmute(
    NHS_Region = normalise_region_name(NHSER26NM),
    NHS_Region_Display = NHSER26NM,
    ONS
  ) %>%
  full_join(practice_region_registered, by = "NHS_Region") %>%
  inner_join(workforce_region_snapshot, by = "NHS_Region") %>%
  filter(GP_FTE > 0) %>%
  mutate(
    Registered_Per_GP_FTE = Registered / GP_FTE,
    ONS_Per_GP_FTE = ONS / GP_FTE,
    Pct_Difference_Per_GP_FTE =
      100 * (Registered_Per_GP_FTE - ONS_Per_GP_FTE) / ONS_Per_GP_FTE
  )

workforce_icb_snapshot <- workforce_snapshot %>%
  inner_join(practice_icb_lookup, by = "Practice_Code") %>%
  group_by(ICB_Key, ICB) %>%
  summarise(GP_FTE = sum(GP_FTE, na.rm = TRUE), .groups = "drop")

practice_icb_registered <- practice_registered_snapshot %>%
  inner_join(practice_icb_lookup, by = "Practice_Code") %>%
  group_by(ICB_Key, ICB) %>%
  summarise(Registered = sum(Registered, na.rm = TRUE), .groups = "drop")

icb_population_snapshot <- icb_snapshot %>%
  transmute(
    ICB_Key = ICB26CD,
    ICB_Population_Name = ICB26NM,
    ONS
  )

patients_per_gp_icb_snapshot <- icb_population_snapshot %>%
  full_join(
    practice_icb_registered %>% rename(ICB_Registered_Name = ICB),
    by = "ICB_Key"
  ) %>%
  inner_join(
    workforce_icb_snapshot %>% rename(ICB_Workforce_Name = ICB),
    by = "ICB_Key"
  ) %>%
  filter(GP_FTE > 0) %>%
  mutate(
    ICB = coalesce(
      ICB_Population_Name,
      ICB_Registered_Name,
      ICB_Workforce_Name
    ),
    Registered_Per_GP_FTE = Registered / GP_FTE,
    ONS_Per_GP_FTE = ONS / GP_FTE
  ) %>%
  select(
    ICB_Key,
    ICB,
    Registered,
    ONS,
    GP_FTE,
    Registered_Per_GP_FTE,
    ONS_Per_GP_FTE
  )

patients_per_gp_national_snapshot <- patients_per_gp_july %>%
  filter(Year == as.integer(format(snapshot_date, "%Y"))) %>%
  select(Source, Patients_Per_GP_FTE) %>%
  pivot_wider(names_from = Source, values_from = Patients_Per_GP_FTE) %>%
  transmute(
    Geography = "England",
    Registered_Per_GP_FTE = `Registered patients`,
    ONS_Per_GP_FTE = `ONS population estimate`
  )

patients_per_gp_region_long <- patients_per_gp_region_snapshot %>%
  select(
    NHS_Region,
    `Registered patients` = Registered_Per_GP_FTE,
    `ONS population` = ONS_Per_GP_FTE
  ) %>%
  pivot_longer(
    -NHS_Region,
    names_to = "Population basis",
    values_to = "Patients_Per_GP_FTE"
  )

region_gp_order <- patients_per_gp_region_snapshot %>%
  arrange(Registered_Per_GP_FTE) %>%
  pull(NHS_Region)

patients_per_gp_region_long <- patients_per_gp_region_long %>%
  mutate(NHS_Region = factor(NHS_Region, levels = region_gp_order))

patients_per_gp_icb_long <- patients_per_gp_icb_snapshot %>%
  select(
    ICB,
    `Registered patients` = Registered_Per_GP_FTE,
    `ONS population` = ONS_Per_GP_FTE
  ) %>%
  pivot_longer(
    -ICB,
    names_to = "Population basis",
    values_to = "Patients_Per_GP_FTE"
  )

icb_gp_order <- patients_per_gp_icb_snapshot %>%
  arrange(Registered_Per_GP_FTE) %>%
  pull(ICB)

patients_per_gp_icb_long <- patients_per_gp_icb_long %>%
  mutate(ICB = factor(ICB, levels = icb_gp_order))
```

## July 2024 national comparison

```{r patients-per-gp-july-2024-bar, fig.width=8, fig.height=5}
patients_per_gp_july_2024 <- patients_per_gp_july %>%
  filter(Period == as.Date("2024-07-01"))

expected_population_sources <- c(
  "ONS population estimate", "Registered patients"
)

if (nrow(patients_per_gp_july_2024) != 2L ||
    !setequal(patients_per_gp_july_2024$Source, expected_population_sources) ||
    any(!is.finite(patients_per_gp_july_2024$GP_FTE)) ||
    any(patients_per_gp_july_2024$GP_FTE <= 0) ||
    n_distinct(patients_per_gp_july_2024$GP_FTE) != 1L ||
    any(!is.finite(patients_per_gp_july_2024$Patients_Per_GP_FTE)) ||
    any(patients_per_gp_july_2024$Patients_Per_GP_FTE < 0)) {
  stop("The July 2024 bar chart requires both population sources and one positive qualified GP FTE total.")
}

patients_per_gp_july_2024 <- patients_per_gp_july_2024 %>%
  mutate(Source = factor(Source, levels = expected_population_sources))

ggplot(
  patients_per_gp_july_2024,
  aes(x = Source, y = Patients_Per_GP_FTE, fill = Source)
) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = scales::comma(Patients_Per_GP_FTE, accuracy = 1)),
    vjust = -0.5,
    size = 4
  ) +
  scale_fill_manual(values = c(
    "ONS population estimate" = "#B2182B",
    "Registered patients" = "#2166AC"
  )) +
  scale_x_discrete(labels = c(
    "ONS population estimate" = "ONS population\nestimate",
    "Registered patients" = "Registered\npatients"
  )) +
  scale_y_continuous(
    limits = c(0, NA),
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Patients per qualified GP FTE",
    subtitle = "England, July 2024",
    x = NULL,
    y = "People per qualified GP FTE",
    caption = paste0(
      "Both bars use the same July 2024 qualified GP total: ",
      scales::comma(patients_per_gp_july_2024$GP_FTE[1], accuracy = 0.1),
      " FTE."
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## July time series

```{r patients-per-gp-july-time-series}
ggplot(
  patients_per_gp_july,
  aes(x = Period, y = Patients_Per_GP_FTE, colour = Source)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    limits = c(0, NA),
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Patients per qualified GP FTE",
    subtitle = "England, July snapshots",
    x = NULL,
    y = "Patients per GP FTE",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

## NHS England regions

```{r patients-per-gp-region-dumbbell, fig.height=7}
gp_region_dumbbell <- patients_per_gp_region_snapshot %>%
  mutate(
    NHS_Region = reorder(NHS_Region, Registered_Per_GP_FTE)
  )

ggplot(
  gp_region_dumbbell,
  aes(y = NHS_Region)
) +
  geom_segment(
    aes(
      x = ONS_Per_GP_FTE,
      xend = Registered_Per_GP_FTE,
      yend = NHS_Region
    ),
    colour = "grey65",
    linewidth = 0.9
  ) +
  geom_point(
    aes(x = Registered_Per_GP_FTE, colour = "Registered patients"),
    size = 3
  ) +
  geom_point(
    aes(x = ONS_Per_GP_FTE, colour = "ONS population"),
    size = 3
  ) +
  geom_text(
    aes(
      x = Registered_Per_GP_FTE,
      label = scales::comma(Registered_Per_GP_FTE, accuracy = 1)
    ),
    hjust = -0.12,
    vjust = -0.7,
    size = 3.4
  ) +
  geom_text(
    aes(
      x = ONS_Per_GP_FTE,
      label = scales::comma(ONS_Per_GP_FTE, accuracy = 1)
    ),
    hjust = -0.12,
    vjust = 1.25,
    size = 3.4
  ) +
  scale_x_continuous(
    labels = scales::label_comma(),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.18))
  ) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population" = "#B2182B"
  )) +
  labs(
    title = "Patients per qualified GP FTE by NHS England region",
    subtitle = format(snapshot_date, "%B %Y"),
    x = "Patients per GP FTE",
    y = NULL,
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r patients-per-gp-region-percentage-difference, fig.height=6}
gp_region_pct_plot <- patients_per_gp_region_snapshot %>%
  mutate(
    NHS_Region = reorder(NHS_Region, Pct_Difference_Per_GP_FTE),
    Direction = if_else(
      Pct_Difference_Per_GP_FTE >= 0,
      "Registered basis higher",
      "Registered basis lower"
    )
  )

ggplot(
  gp_region_pct_plot,
  aes(
    x = Pct_Difference_Per_GP_FTE,
    y = NHS_Region,
    fill = Direction
  )
) +
  geom_vline(xintercept = 0, colour = "grey55") +
  geom_col(width = 0.7) +
  geom_text(
    aes(
      label = sprintf("%.1f%%", Pct_Difference_Per_GP_FTE),
      hjust = if_else(Pct_Difference_Per_GP_FTE >= 0, -0.12, 1.12)
    ),
    size = 3.6
  ) +
  scale_x_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0.12, 0.15))  ) +
  scale_fill_manual(values = c(
    "Registered basis higher" = "#2166AC",
    "Registered basis lower" = "#B2182B"
  ), guide = "none") +
  labs(
    title = "Percentage difference in patients per qualified GP FTE by NHS England region",
    subtitle = paste0(
      "Registered-patient basis compared with ONS population, ",
      format(snapshot_date, "%B %Y")
    ),
    x = "Percentage difference vs ONS basis",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Integrated care boards

```{r patients-per-gp-icb-dumbbell, fig.height=13}
gp_icb_dumbbell <- patients_per_gp_icb_snapshot %>%
  mutate(
    ICB = reorder(ICB, Registered_Per_GP_FTE)
  )

ggplot(
  gp_icb_dumbbell,
  aes(y = ICB)
) +
  geom_segment(
    aes(
      x = ONS_Per_GP_FTE,
      xend = Registered_Per_GP_FTE,
      yend = ICB
    ),
    colour = "grey65",
    linewidth = 0.8
  ) +
  geom_point(
    aes(x = Registered_Per_GP_FTE, colour = "Registered patients"),
    size = 2.5
  ) +
  geom_point(
    aes(x = ONS_Per_GP_FTE, colour = "ONS population"),
    size = 2.5
  ) +
  geom_text(
    aes(
      x = Registered_Per_GP_FTE,
      label = scales::comma(Registered_Per_GP_FTE, accuracy = 1)
    ),
    hjust = -0.10,
    vjust = -0.7,
    size = 2.8
  ) +
  geom_text(
    aes(
      x = ONS_Per_GP_FTE,
      label = scales::comma(ONS_Per_GP_FTE, accuracy = 1)
    ),
    hjust = -0.10,
    vjust = 1.25,
    size = 2.8
  ) +
  scale_x_continuous(
    labels = scales::label_comma(),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.20))
  ) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population" = "#B2182B"
  )) +
  labs(
    title = "Patients per qualified GP FTE by NHS integrated care board",
    subtitle = format(snapshot_date, "%B %Y"),
    x = "Patients per GP FTE",
    y = NULL,
    colour = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r patients-per-gp-table, results='asis'}
patients_per_gp_national_snapshot %>%
  transmute(
    Geography,
    `Registered patients per GP FTE` =
      scales::comma(Registered_Per_GP_FTE, accuracy = 1),
    `ONS population per GP FTE` =
      scales::comma(ONS_Per_GP_FTE, accuracy = 1)
  ) %>%
  report_table(caption = "National patients per GP FTE comparison")

patients_per_gp_region_snapshot %>%
  arrange(desc(Registered_Per_GP_FTE)) %>%
  transmute(
    `NHS England region` = NHS_Region,
    `Registered patients per GP FTE` =
      scales::comma(Registered_Per_GP_FTE, accuracy = 1),
    `ONS population per GP FTE` =
      scales::comma(ONS_Per_GP_FTE, accuracy = 1)
  ) %>%
  report_table(caption = "Regional patients per GP FTE comparison")

patients_per_gp_icb_snapshot %>%
  arrange(desc(Registered_Per_GP_FTE)) %>%
  transmute(
    ICB,
    `Registered patients per GP FTE` =
      scales::comma(Registered_Per_GP_FTE, accuracy = 1),
    `ONS population per GP FTE` =
      scales::comma(ONS_Per_GP_FTE, accuracy = 1)
  ) %>%
  report_table(caption = "ICB patients per GP FTE comparison")
```

# Potential overall payment reduction

## Method

This section models the money removed from payments if registered lists were reduced towards the ONS population and each removed registration carried its payment with it. For each geography, the annual payment per registered patient is calculated from its observed total payments and its registered-patient count:

\[
\text{Payment per registered patient} = \frac{\text{Current payments}}{\text{Registered patients}}
\]

\[
\text{Registrations removed} = \max(\text{Registered patients} - \text{ONS residents},\ 0)
\]

\[
\text{Potential loss} = \text{Registrations removed} \times \text{Payment per registered patient}
\]

\[
\text{Remaining payments} = \text{Current payments} - \text{Potential loss}
\]

For example, a payment rate of **£165 per registered patient** and a modelled reduction of **1,000 registrations** produces a **£165,000 loss**, subtracted from the current total. The report uses the actual calculated rate for each geography, rather than fixing every rate at £165. Where the ONS population equals or exceeds registrations, this loss-only scenario removes no registrations and shows no payment reduction.

The scenario assumes that all payments fall proportionately with the unweighted registered list. Actual payments include categories such as premises, QOF and incentive schemes, so [this is an illustrative funding scenario rather than a prediction of contractual losses](https://digital.nhs.uk/data-and-information/publications/statistical/nhs-payments-to-general-practice/england-2024-25). It also assumes that the entire positive registration-versus-ONS difference could be removed; the difference does not identify individual ghost patients. Practice counts follow practice location while ONS counts follow residence, so cross-boundary registration can contribute to the difference.

National and local estimates use their own observed payment rates and population differences. Regional or ICB losses therefore need not sum to the independently calculated national loss, especially where positive and negative population differences offset. Local figures include records assigned to the relevant geography; unassigned payment and registration amounts are disclosed below. All population counts retain the selected 2024 snapshot, with ICB results grouped on April 2026 geography.

```{r payment-scenario-prepare, include=FALSE, results='asis'}
source("13_Model_Payment_Reduction.R")
payment_metric_name <- "Total NHS Payments to General Practice Minus Deductions"

payment_data <- finance_datasets %>%
  mutate(
    Effective_Snapshot_Date = as.Date(Effective_Snapshot_Date),
    Metric_Value = as.numeric(Metric_Value),
    Practice_Code = as.character(Practice_Code)
  ) %>%
  filter(Metric == payment_metric_name)

if (nrow(payment_data) == 0L ||
    any(is.na(payment_data$Effective_Snapshot_Date))) {
  stop("No valid payment snapshot is available for the selected metric.")
}
payment_snapshot_date <- max(payment_data$Effective_Snapshot_Date)
payment_current <- payment_data %>%
  filter(Effective_Snapshot_Date == payment_snapshot_date)
if (any(!is.finite(payment_current$Metric_Value))) {
  stop("The selected payment snapshot contains missing or non-finite amounts.")
}
payment_by_practice <- payment_current %>%
  group_by(Practice_Code) %>%
  summarise(Total_Payments = sum(Metric_Value), .groups = "drop")

national_payment_scenario <- model_payment_reduction(tibble::tibble(
  Geography = "England",
  Total_Payments = sum(payment_by_practice$Total_Payments),
  Registered = national_snapshot$Registered,
  ONS = national_snapshot$ONS
))

regional_payment_scenario <- payment_by_practice %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(Total_Payments = sum(Total_Payments), .groups = "drop") %>%
  left_join(practice_region_registered, by = "NHS_Region") %>%
  left_join(
    region_snapshot %>%
      transmute(NHS_Region = normalise_region_name(NHSER26NM), ONS),
    by = "NHS_Region"
  ) %>%
  model_payment_reduction()

icb_payment_scenario <- payment_by_practice %>%
  inner_join(practice_icb_lookup, by = "Practice_Code") %>%
  group_by(ICB_Key, ICB) %>%
  summarise(Total_Payments = sum(Total_Payments), .groups = "drop") %>%
  rename(ICB_Payment_Name = ICB) %>%
  left_join(
    practice_icb_registered %>% rename(ICB_Registered_Name = ICB),
    by = "ICB_Key"
  ) %>%
  left_join(icb_population_snapshot, by = "ICB_Key") %>%
  mutate(ICB = coalesce(ICB_Population_Name, ICB_Registered_Name, ICB_Payment_Name)) %>%
  model_payment_reduction()

payment_unassigned_region <- payment_by_practice %>%
  anti_join(practice_region_lookup, by = "Practice_Code") %>%
  summarise(Amount = sum(Total_Payments)) %>% pull(Amount)
payment_unassigned_icb <- payment_by_practice %>%
  anti_join(practice_icb_lookup, by = "Practice_Code") %>%
  summarise(Amount = sum(Total_Payments)) %>% pull(Amount)
registered_unassigned_region <- national_snapshot$Registered -
  sum(practice_region_registered$Registered)
registered_unassigned_icb <- national_snapshot$Registered -
  sum(practice_icb_registered$Registered)
if (registered_unassigned_region < -0.5 || registered_unassigned_icb < -0.5) {
  stop("Geographic registration totals exceed the national practice total.")
}

payment_currency <- function(x, accuracy = 1) {
  scales::dollar(x, prefix = gbp_symbol, accuracy = accuracy)
}
payment_loss_chart <- function(data, geography, title) {
  ggplot(data, aes(
    x = Potential_Reduction,
    y = reorder(.data[[geography]], Potential_Reduction)
  )) +
    geom_col(fill = "#2166AC", width = 0.65) +
    geom_text(
      aes(label = paste0(
        gbp_symbol, scales::number(Potential_Reduction / 1e6, accuracy = 0.1),
        "m (", sprintf("%.1f%%", Reduction_Pct), ")"
      )),
      hjust = -0.08, size = 3.1
    ) +
    scale_x_continuous(
      labels = scales::label_currency(prefix = gbp_symbol, scale = 1e-6, suffix = "m"),
      limits = c(0, NA), expand = expansion(mult = c(0, 0.25))
    ) +
    labs(
      title = title,
      subtitle = paste("Payment snapshot:", format(payment_snapshot_date, "%d %b %Y"),
                       "| Population:", format(snapshot_date, "%B %Y")),
      x = "Modelled payment loss", y = NULL,
      caption = "Labels show money lost and percentage of current payments."
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
}
payment_loss_table <- function(data, geography, caption) {
  data %>%
    arrange(desc(Potential_Reduction)) %>%
    transmute(
      Geography = .data[[geography]],
      `Current (£)` = scales::comma(Total_Payments, accuracy = 1),
      `Rate (£/patient)` = scales::number(Payment_Per_Registered, accuracy = 0.01),
      `List reduction (n)` = scales::comma(Registrations_Removed),
      `Loss (£)` = scales::comma(Potential_Reduction, accuracy = 1),
      `Remaining (£)` = scales::comma(Remaining_Payments, accuracy = 1)
    ) %>%
    report_table(caption = caption)
}
```

The payment measure is **Total NHS Payments to General Practice Minus Deductions**, using the latest available payment snapshot (`r format(payment_snapshot_date, "%d %B %Y")`). Population counts refer to `r format(snapshot_date, "%B %Y")`; the payment period and population snapshot are therefore not identical. Payment totals relate to the annual payment records, not one month's expenditure.

Amounts without a region assignment are `r payment_currency(payment_unassigned_region)`, alongside `r scales::comma(registered_unassigned_region)` registrations without a region assignment. Amounts without an ICB assignment are `r payment_currency(payment_unassigned_icb)`, alongside `r scales::comma(registered_unassigned_icb)` registrations without an ICB assignment. These national records are outside the corresponding local estimates.

## National loss

At the national calculated rate of **`r payment_currency(national_payment_scenario$Payment_Per_Registered, accuracy = 0.01)` per registered patient**, a modelled removal of **`r scales::comma(national_payment_scenario$Registrations_Removed)` registrations** produces a potential loss of **`r payment_currency(national_payment_scenario$Potential_Reduction)`** (`r sprintf("%.1f%%", national_payment_scenario$Reduction_Pct)` of current payments). This is subtracted from **`r payment_currency(national_payment_scenario$Total_Payments)`**, leaving **`r payment_currency(national_payment_scenario$Remaining_Payments)`**.

```{r national-payment-loss-chart, fig.height=3.5}
payment_loss_chart(national_payment_scenario, "Geography", "Potential national payment loss")
```

```{r national-payment-loss-table, results='asis'}
payment_loss_table(national_payment_scenario, "Geography", "National payment loss scenario")
```

## Loss by NHS England region

```{r regional-payment-loss-chart, fig.height=7}
payment_loss_chart(regional_payment_scenario, "NHS_Region",
                   "Potential payment loss by NHS England region")
```

```{r regional-payment-loss-table, results='asis'}
payment_loss_table(regional_payment_scenario, "NHS_Region", "Regional payment loss scenarios")
```

## Loss by integrated care board

```{r icb-payment-loss-chart, fig.height=13}
payment_loss_chart(icb_payment_scenario, "ICB",
                   "Potential payment loss by ICB (April 2026 geography)")
```

```{r icb-payment-loss-table, results='asis'}
payment_loss_table(icb_payment_scenario, "ICB", "ICB payment loss scenarios")
```

## Data sources

- GP registered-patient national, age, and sex data: Demography.No_Of_Patients_Regd_At_GP_Practice_Single_Age1.
- GP registered-patient LSOA data: Demography.No_Of_Patients_Regd_At_GP_Practice_LSOA_2021_Level1. The source pull includes all residence codes for patients registered with English GP practices; all LSOA-derived report analyses use an explicit English E01... subset, with excluded non-English registered patients reported below the first LSOA chart.
- ONS population estimates: Demography.ONS_Population_Estimates_For_LSOAs_By_Year_Of_Age1, restricted to English LSOAs.
- LSOA IMD: latest snapshot in Demography.Index_Of_Multiple_Deprivation_By_LSOA1.
- GP workforce: NHS_Workforce.Practice_Level_Census_Data1, summing Measure_Value where Measure is TOTAL_GP_SEN_PTNR_FTE, TOTAL_GP_PTNR_PROV_FTE, TOTAL_GP_SAL_BY_PRAC_FTE, or TOTAL_GP_SAL_BY_OTH_FTE.
- GP payments: NHS_Payments_To_GP_Practices.Figures1 via 03_Payments_To_GP_Practice.R.
- LSOA-to-ICB lookup: [official ONS April 2026 LSOA21-to-SICBL26-to-ICB26-to-NHSER26 lookup](https://www.data.gov.uk/dataset/1d16e489-1dc8-443e-9e1c-15de4f961c6d/lsoa-2021-to-sicbl-to-icb-to-nhser-to-lad-april-2026-lookup-in-en), applied to the July 2024 source counts.
- Practice-to-region lookup: payments2425.csv.
- Practice-to-ICB mapping: retained Sub ICB codes in payments2425.csv mapped to April 2026 ICBs. Frimley transfers use the NHS May 2026 practice mapping; nine historical practices absent from that mapping use their recorded practice postcode in the ONS May 2026 directory. The postcode assignments are geographic estimates for closed practices, not confirmed commissioning relationships. See data/README.md for sources and methods.
