---
title: "Registered patients compared with ONS population estimates"
subtitle: "Focused national and LSOA-level analysis"
output:
  word_document: default
  html_document:
    toc: true
    toc_float: true
params:
  start_date: "2015-07-01"
  end_date: "2026-08-31"
  snapshot_date: "2024-07-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  dpi = 150
)

required_packages <- c(
  "DBI", "odbc", "dplyr", "tidyr", "ggplot2", "scales",
  "knitr", "httr", "jsonlite", "readr", "glue", "rlang", "tibble"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the following R packages before rendering: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(DBI)
library(odbc)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(knitr)

start_date <- as.Date(params$start_date)
end_date <- as.Date(params$end_date)
snapshot_date <- as.Date(params$snapshot_date)

stopifnot(
  !is.na(start_date),
  !is.na(end_date),
  !is.na(snapshot_date),
  start_date <= end_date
)

pct_difference <- function(registered, ons) {
  ifelse(
    is.na(ons) | ons == 0,
    NA_real_,
    100 * (registered - ons) / ons
  )
}

comparison_bar <- function(data, category, title, subtitle, y_title = NULL) {
  category_quo <- rlang::enquo(category)

  ggplot(
    data,
    aes(
      x = Pct_Difference,
      y = !!category_quo,
      fill = Pct_Difference
    )
  ) +
    geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.5) +
    geom_col(width = 0.7) +
    geom_text(
      aes(label = sprintf("%.1f%%", Pct_Difference)),
      hjust = ifelse(data$Pct_Difference >= 0, -0.12, 1.12),
      size = 3.7
    ) +
    scale_x_continuous(
      labels = function(x) paste0(x, "%"),
      expand = expansion(mult = c(0.16, 0.18))
    ) +
    scale_fill_gradient2(
      low = "#B2182B",
      mid = "#F2F2F2",
      high = "#2166AC",
      midpoint = 0,
      guide = "none"
    ) +
    coord_cartesian(clip = "off") +
    labs(
      title = title,
      subtitle = subtitle,
      x = "Percentage difference from ONS",
      y = y_title,
      caption = "Percentage difference = (registered patients − ONS estimate) / ONS estimate × 100."
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = margin(5.5, 30, 5.5, 5.5)
    )
}
```

# Scope and interpretation

This report compares GP registered-patient counts with ONS population estimates for England. All snapshot charts use `r format(snapshot_date, "%B %Y")` and report the percentage difference relative to ONS:

\[
100 \times \frac{\text{registered patients} - \text{ONS population}}{\text{ONS population}}
\]

Positive values indicate that registered-patient counts exceed ONS estimates. Negative values indicate that registered-patient counts are lower. Sex-specific charts include female and male records only. The NHS-region lookup represents boundaries as at April 2026.

```{r connect-and-load-data, include=FALSE}
source("01_connect_to_facts_and_dimensions.R")
source("08_Get_Focused_Patient_List_Data.R")
source("09_Get_LSOA_NHS_Region_Lookup.R")
DBI::dbDisconnect(con)
```

```{r prepare-data, include=FALSE}
national_time_series <- bind_rows(
  national_registered,
  national_ons
)

lsoa_snapshot <- full_join(
  lsoa_registered_snapshot,
  lsoa_ons_snapshot,
  by = "LSOA_Code"
) %>%
  mutate(
    Registered = replace_na(Registered, 0),
    ONS = replace_na(ONS, 0),
    Pct_Difference = pct_difference(Registered, ONS)
  )

demographic_snapshot <- full_join(
  registered_demographic_snapshot,
  ons_demographic_snapshot,
  by = c("Sex", "Age_Band")
) %>%
  mutate(
    Registered = replace_na(Registered, 0),
    ONS = replace_na(ONS, 0),
    Pct_Difference = pct_difference(Registered, ONS)
  )

age_levels <- c(
  "0-9", "10-19", "20-29", "30-39", "40-49",
  "50-59", "60-69", "70-79", "80-89", "90+"
)

demographic_snapshot <- demographic_snapshot %>%
  mutate(
    Age_Band = factor(Age_Band, levels = age_levels),
    Sex = factor(Sex, levels = c("Female", "Male"))
  )
```

# National total over time

```{r national-time-series}
ggplot(
  national_time_series,
  aes(
    x = Period,
    y = Population,
    colour = Source,
    group = Source
  )
) +
  geom_line(linewidth = 1.15) +
  geom_point(
    data = national_time_series %>%
      filter(Source == "ONS population estimate"),
    size = 2.2
  ) +
  scale_colour_manual(
    values = c(
      "Registered patients" = "#2166AC",
      "ONS population estimate" = "#B2182B"
    )
  ) +
  scale_x_date(
    limits = c(start_date, end_date),
    date_breaks = "1 year",
    date_labels = "%b\n%Y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "m")
  ) +
  labs(
    title = "England population: registered patients and ONS estimates",
    subtitle = paste0(
      format(start_date, "%B %Y"), " to ",
      format(end_date, "%B %Y"), "; available snapshots"
    ),
    x = NULL,
    y = "Population",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r latest-data-table}
latest_dates <- national_time_series %>%
  group_by(Source) %>%
  summarise(
    `Latest available snapshot` = max(Period, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  latest_dates,
  caption = "Latest available source dates within the requested period"
)
```

# National snapshot

```{r national-snapshot}
national_snapshot <- lsoa_snapshot %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE)
  ) %>%
  mutate(
    Geography = "England",
    Pct_Difference = pct_difference(Registered, ONS)
  )

comparison_bar(
  national_snapshot,
  Geography,
  title = "National difference between registered patients and ONS estimate",
  subtitle = format(snapshot_date, "%B %Y"),
  y_title = NULL
)
```

```{r national-snapshot-table}
national_snapshot %>%
  transmute(
    Geography,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Difference` = scales::comma(Registered - ONS),
    `Percentage difference from ONS` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  knitr::kable(
    caption = paste("National comparison:", format(snapshot_date, "%B %Y"))
  )
```

# Difference by age and sex

```{r age-sex-bars, fig.height=6}
ggplot(
  demographic_snapshot,
  aes(
    x = Age_Band,
    y = Pct_Difference,
    fill = Sex
  )
) +
  geom_hline(yintercept = 0, colour = "grey55", linewidth = 0.5) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", Pct_Difference)),
    position = position_dodge(width = 0.8),
    vjust = ifelse(demographic_snapshot$Pct_Difference >= 0, -0.25, 1.15),
    size = 3,
    angle = 90
  ) +
  scale_fill_manual(
    values = c("Female" = "#CC79A7", "Male" = "#0072B2")
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0.12, 0.18))
  ) +
  labs(
    title = "Difference from ONS by age and sex",
    subtitle = format(snapshot_date, "%B %Y"),
    x = "Age band",
    y = "Percentage difference from ONS",
    fill = "Sex",
    caption = "Female and male records only."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Difference by age

```{r age-bars, fig.height=6}
age_snapshot <- demographic_snapshot %>%
  group_by(Age_Band) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Pct_Difference = pct_difference(Registered, ONS))

comparison_bar(
  age_snapshot,
  Age_Band,
  title = "Difference from ONS by age",
  subtitle = format(snapshot_date, "%B %Y"),
  y_title = "Age band"
)
```

# Difference by sex

```{r sex-bars}
sex_snapshot <- demographic_snapshot %>%
  group_by(Sex) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Pct_Difference = pct_difference(Registered, ONS))

comparison_bar(
  sex_snapshot,
  Sex,
  title = "Difference from ONS by sex",
  subtitle = format(snapshot_date, "%B %Y"),
  y_title = "Sex"
)
```

# Difference by LSOA-level IMD decile

```{r imd-bars, fig.height=6}
imd_snapshot <- lsoa_snapshot %>%
  inner_join(
    imd_lookup %>%
      mutate(IMD_Decile = as.integer(IMD_Decile)),
    by = "LSOA_Code"
  ) %>%
  filter(IMD_Decile %in% 1:10) %>%
  group_by(IMD_Decile) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Pct_Difference = pct_difference(Registered, ONS),
    IMD_Label = factor(
      IMD_Decile,
      levels = 10:1,
      labels = c(
        "10 (Least deprived)", "9", "8", "7", "6",
        "5", "4", "3", "2", "1 (Most deprived)"
      )
    )
  )

comparison_bar(
  imd_snapshot,
  IMD_Label,
  title = "Difference from ONS by LSOA-level IMD decile",
  subtitle = paste(
    format(snapshot_date, "%B %Y"),
    "using the latest available LSOA IMD lookup"
  ),
  y_title = "IMD decile"
)
```

# Difference by NHS England region

```{r nhs-region-bars, fig.height=6}
region_snapshot <- lsoa_snapshot %>%
  inner_join(
    nhser_lookup %>%
      select(LSOA21CD, NHSER26NM),
    by = c("LSOA_Code" = "LSOA21CD")
  ) %>%
  filter(!is.na(NHSER26NM)) %>%
  group_by(NHSER26NM) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Pct_Difference = pct_difference(Registered, ONS),
    NHSER26NM = reorder(NHSER26NM, Pct_Difference)
  )

comparison_bar(
  region_snapshot,
  NHSER26NM,
  title = "Difference from ONS by NHS England region",
  subtitle = paste0(
    format(snapshot_date, "%B %Y"),
    "; LSOA mapped to April 2026 NHS regions"
  ),
  y_title = "NHS England region"
)
```

```{r lookup-coverage}
coverage <- tibble::tibble(
  Lookup = c("LSOA IMD decile", "NHS England region"),
  `Mapped registered population` = c(
    sum(imd_snapshot$Registered, na.rm = TRUE),
    sum(region_snapshot$Registered, na.rm = TRUE)
  ),
  `National registered population` = national_snapshot$Registered
) %>%
  mutate(
    `Mapping coverage` = scales::percent(
      `Mapped registered population` / `National registered population`,
      accuracy = 0.1
    ),
    `Mapped registered population` = scales::comma(`Mapped registered population`),
    `National registered population` = scales::comma(`National registered population`)
  )

knitr::kable(
  coverage,
  caption = "LSOA lookup coverage for the snapshot analyses"
)
```

## Data sources

- GP registered-patient data: `Demography.No_Of_Patients_Regd_At_GP_Practice_LSOA_2021_Level1`.
- ONS population estimates: `Demography.ONS_Population_Estimates_For_LSOAs_By_Year_Of_Age1`.
- LSOA IMD: latest snapshot in `Demography.Index_Of_Multiple_Deprivation_By_LSOA1`.
- LSOA-to-NHS-region mapping: ONS, *LSOA (2021) to SICBL to ICB to NHSER to LAD (April 2026) Lookup in England*.
