---
title: "Registered patients compared with ONS population estimates"
subtitle: "Focused national and LSOA-level analysis"
output:
  word_document: default
  html_document:
    toc: true
    toc_float: true
params:
  start_date: "2015-07-01"
  end_date: "2026-08-31"
  snapshot_date: "2024-07-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  dpi = 150
)

required_packages <- c(
  "DBI", "odbc", "dplyr", "tidyr", "ggplot2", "scales",
  "knitr", "httr", "jsonlite", "readr", "glue", "rlang", "tibble",
  "tidyverse"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the following R packages before rendering: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(DBI)
library(odbc)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(knitr)

start_date <- as.Date(params$start_date)
end_date <- as.Date(params$end_date)
snapshot_date <- as.Date(params$snapshot_date)

stopifnot(
  !is.na(start_date),
  !is.na(end_date),
  !is.na(snapshot_date),
  start_date <= end_date
)

pct_difference <- function(registered, ons) {
  ifelse(
    is.na(ons) | ons == 0,
    NA_real_,
    100 * (registered - ons) / ons
  )
}

comparison_bar <- function(data, category, title, subtitle, y_title = NULL) {
  category_quo <- rlang::enquo(category)

  ggplot(
    data,
    aes(
      x = Pct_Difference,
      y = !!category_quo,
      fill = Pct_Difference
    )
  ) +
    geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.5) +
    geom_col(width = 0.7) +
    geom_text(
      aes(label = sprintf("%.1f%%", Pct_Difference)),
      hjust = ifelse(data$Pct_Difference >= 0, -0.12, 1.12),
      size = 3.7
    ) +
    scale_x_continuous(
      labels = function(x) paste0(x, "%"),
      expand = expansion(mult = c(0.16, 0.18))
    ) +
    scale_fill_gradient2(
      low = "#B2182B",
      mid = "#F2F2F2",
      high = "#2166AC",
      midpoint = 0,
      guide = "none"
    ) +
    coord_cartesian(clip = "off") +
    labs(
      title = title,
      subtitle = subtitle,
      x = "Percentage difference from ONS",
      y = y_title,
      caption = "Percentage difference = (registered patients − ONS estimate) / ONS estimate × 100."
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = margin(5.5, 30, 5.5, 5.5)
    )
}
```

# Scope and interpretation

This report compares GP registered-patient counts with ONS population estimates for England from July 2015 onwards. Percentage difference is defined as:

\[
100 \times \frac{\text{registered patients} - \text{ONS population}}{\text{ONS population}}
\]

Positive values mean registered-patient counts exceed ONS estimates. Snapshot analyses use `r format(snapshot_date, "%B %Y")`.

```{r connect-and-load-data, include=FALSE}
source("01_connect_to_facts_and_dimensions.R")
source("03_Payments_To_GP_Practice.R")
source("08_Get_Focused_Patient_List_Data.R")
source("09_Get_LSOA_NHS_Region_Lookup.R")
source("10_Get_Practice_Region_Lookup.R")
source("11_Get_NHS_Boundaries.R")
DBI::dbDisconnect(con)
```

```{r prepare-analysis-data, include=FALSE}
age_levels <- c(
  "0-9", "10-19", "20-29", "30-39", "40-49",
  "50-59", "60-69", "70-79", "80-89", "90+"
)

national_time_series <- bind_rows(national_registered, national_ons)

demographic_time_series <- full_join(
  practice_registered_demographic_time_series,
  ons_demographic_time_series,
  by = c("Period", "Sex", "Age_Band")
) %>%
  mutate(
    Age_Band = factor(Age_Band, levels = age_levels),
    Sex = factor(Sex, levels = c("Female", "Male")),
    Pct_Difference = pct_difference(Registered, ONS)
  )

sum_if_available <- function(x) {
  if (all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
}

age_time_series <- demographic_time_series %>%
  group_by(Period, Age_Band) %>%
  summarise(
    Registered = sum_if_available(Registered),
    ONS = sum_if_available(ONS),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Registered, ONS),
    names_to = "Source",
    values_to = "Population"
  ) %>%
  mutate(
    Source = recode(
      Source,
      Registered = "Registered patients",
      ONS = "ONS population estimate"
    )
  )

sex_time_series <- demographic_time_series %>%
  group_by(Period, Sex) %>%
  summarise(
    Registered = sum_if_available(Registered),
    ONS = sum_if_available(ONS),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Registered, ONS),
    names_to = "Source",
    values_to = "Population"
  ) %>%
  mutate(
    Source = recode(
      Source,
      Registered = "Registered patients",
      ONS = "ONS population estimate"
    )
  )

national_snapshot <- national_time_series %>%
  filter(Period == snapshot_date) %>%
  select(Source, Population) %>%
  pivot_wider(names_from = Source, values_from = Population) %>%
  transmute(
    Geography = "England",
    Registered = `Registered patients`,
    ONS = `ONS population estimate`,
    Pct_Difference = pct_difference(Registered, ONS)
  )

age_sex_snapshot <- demographic_time_series %>%
  filter(Period == snapshot_date)

age_snapshot <- age_sex_snapshot %>%
  group_by(Age_Band) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Pct_Difference = pct_difference(Registered, ONS))

sex_snapshot <- age_sex_snapshot %>%
  group_by(Sex) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Pct_Difference = pct_difference(Registered, ONS))

lsoa_july <- full_join(
  lsoa_registered_july,
  lsoa_ons_july,
  by = c("LSOA_Code", "Period")
) %>%
  mutate(
    Registered = replace_na(Registered, 0),
    ONS = replace_na(ONS, 0),
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

lsoa_july_geography <- lsoa_july %>%
  inner_join(nhser_lookup, by = c("LSOA_Code" = "LSOA21CD"))

imd_july <- lsoa_july %>%
  inner_join(
    imd_lookup %>% mutate(IMD_Decile = as.integer(IMD_Decile)),
    by = "LSOA_Code"
  ) %>%
  filter(IMD_Decile %in% 1:10) %>%
  group_by(Period, IMD_Decile) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS),
    IMD_Label = factor(
      IMD_Decile,
      levels = 10:1,
      labels = c(
        "10 (Least deprived)", "9", "8", "7", "6",
        "5", "4", "3", "2", "1 (Most deprived)"
      )
    )
  )

region_july <- lsoa_july_geography %>%
  group_by(Period, NHSER26CD, NHSER26NM) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

icb_july <- lsoa_july_geography %>%
  group_by(Period, ICB26CD, ICB26NM) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

imd_snapshot <- imd_july %>% filter(Period == snapshot_date)
region_snapshot <- region_july %>% filter(Period == snapshot_date)
icb_snapshot <- icb_july %>% filter(Period == snapshot_date)

region_raw_change <- region_july %>%
  group_by(NHSER26CD, NHSER26NM) %>%
  arrange(Period, .by_group = TRUE) %>%
  mutate(
    Raw_Difference_Pct_Change = 100 *
      (Raw_Difference - lag(Raw_Difference)) / abs(lag(Raw_Difference))
  ) %>%
  ungroup() %>%
  filter(
    Period == snapshot_date,
    is.finite(Raw_Difference_Pct_Change)
  )

icb_raw_change <- icb_july %>%
  group_by(ICB26CD, ICB26NM) %>%
  arrange(Period, .by_group = TRUE) %>%
  mutate(
    Raw_Difference_Pct_Change = 100 *
      (Raw_Difference - lag(Raw_Difference)) / abs(lag(Raw_Difference))
  ) %>%
  ungroup() %>%
  filter(
    Period == snapshot_date,
    is.finite(Raw_Difference_Pct_Change)
  )
```

# National population over time

```{r national-population-time-series}
ggplot(
  national_time_series,
  aes(x = Period, y = Population, colour = Source)
) +
  geom_line(linewidth = 1.1) +
  geom_point(
    data = national_time_series %>%
      filter(Source == "ONS population estimate"),
    size = 2
  ) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(
    limits = c(start_date, end_date),
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "m")
  ) +
  labs(
    title = "England population: registered patients and ONS estimates",
    subtitle = "July 2015 onwards; available source snapshots",
    x = NULL,
    y = "Population",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

# Age over time

```{r age-time-series, fig.height=10}
ggplot(
  age_time_series,
  aes(x = Period, y = Population, colour = Source)
) +
  geom_line(linewidth = 0.8) +
  geom_point(
    data = age_time_series %>%
      filter(Source == "ONS population estimate"),
    size = 1.3
  ) +
  facet_wrap(~ Age_Band, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "m")) +
  labs(
    title = "Population by age over time",
    subtitle = "Registered patients and ONS estimates",
    x = NULL,
    y = "Population",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

# Sex over time

```{r sex-time-series, fig.height=7}
ggplot(
  sex_time_series,
  aes(x = Period, y = Population, colour = Source)
) +
  geom_line(linewidth = 1) +
  geom_point(
    data = sex_time_series %>%
      filter(Source == "ONS population estimate"),
    size = 1.8
  ) +
  facet_wrap(~ Sex, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "m")) +
  labs(
    title = "Population by sex over time",
    subtitle = "Registered patients and ONS estimates",
    x = NULL,
    y = "Population",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

# July 2024 national snapshots

## National total

```{r national-snapshot-chart}
comparison_bar(
  national_snapshot,
  Geography,
  "National percentage difference",
  format(snapshot_date, "%B %Y")
)
```

```{r national-snapshot-table}
national_snapshot %>%
  transmute(
    Geography,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  knitr::kable(caption = "National population comparison")
```

## Age

```{r age-snapshot-chart, fig.height=6}
comparison_bar(
  age_snapshot,
  Age_Band,
  "Percentage difference by age",
  format(snapshot_date, "%B %Y"),
  "Age band"
)
```

```{r age-snapshot-table}
age_snapshot %>%
  transmute(
    `Age band` = Age_Band,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  knitr::kable(caption = "Population comparison by age")
```

## Sex

```{r sex-snapshot-chart}
comparison_bar(
  sex_snapshot,
  Sex,
  "Percentage difference by sex",
  format(snapshot_date, "%B %Y"),
  "Sex"
)
```

```{r sex-snapshot-table}
sex_snapshot %>%
  transmute(
    Sex,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  knitr::kable(caption = "Population comparison by sex")
```

## Age and sex

```{r age-sex-snapshot-chart, fig.height=7}
ggplot(
  age_sex_snapshot,
  aes(x = Age_Band, y = Pct_Difference, fill = Sex)
) +
  geom_hline(yintercept = 0, colour = "grey55") +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_fill_manual(values = c(Female = "#CC79A7", Male = "#0072B2")) +
  labs(
    title = "Percentage difference by age and sex",
    subtitle = format(snapshot_date, "%B %Y"),
    x = "Age band",
    y = "Percentage difference from ONS",
    fill = "Sex"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r age-sex-snapshot-table}
age_sex_snapshot %>%
  arrange(Age_Band, Sex) %>%
  transmute(
    `Age band` = Age_Band,
    Sex,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  knitr::kable(caption = "Population comparison by age and sex")
```

# IMD deciles using LSOA data

```{r imd-time-series, fig.height=7}
ggplot(
  imd_july,
  aes(x = Period, y = Pct_Difference, colour = factor(IMD_Decile))
) +
  geom_hline(yintercept = 0, colour = "grey65") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Percentage difference by IMD decile over time",
    subtitle = "July snapshots where LSOA codes map to the latest IMD lookup",
    x = NULL,
    y = "Percentage difference from ONS",
    colour = "IMD decile"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

```{r imd-snapshot-chart, fig.height=6}
comparison_bar(
  imd_snapshot,
  IMD_Label,
  "Percentage difference by LSOA-level IMD decile",
  format(snapshot_date, "%B %Y"),
  "IMD decile"
)
```

```{r imd-snapshot-table}
imd_snapshot %>%
  arrange(IMD_Decile) %>%
  transmute(
    `IMD decile` = IMD_Label,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  knitr::kable(caption = "Population comparison by IMD decile")
```

# NHS England regions

```{r region-snapshot-chart, fig.height=6}
region_snapshot_plot <- region_snapshot %>%
  mutate(NHSER26NM = reorder(NHSER26NM, Pct_Difference))

comparison_bar(
  region_snapshot_plot,
  NHSER26NM,
  "Percentage difference by NHS England region",
  format(snapshot_date, "%B %Y"),
  "NHS England region"
)
```

```{r region-snapshot-table}
region_snapshot %>%
  arrange(desc(Pct_Difference)) %>%
  transmute(
    `NHS England region` = NHSER26NM,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  knitr::kable(caption = "Population comparison by NHS England region")
```

# NHS integrated care boards

```{r icb-snapshot-chart, fig.height=12}
icb_snapshot_plot <- icb_snapshot %>%
  mutate(ICB26NM = reorder(ICB26NM, Pct_Difference))

comparison_bar(
  icb_snapshot_plot,
  ICB26NM,
  "Percentage difference by NHS integrated care board",
  format(snapshot_date, "%B %Y"),
  "Integrated care board"
)
```

```{r icb-snapshot-table}
icb_snapshot %>%
  arrange(desc(Pct_Difference)) %>%
  transmute(
    ICB = ICB26NM,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  knitr::kable(caption = "Population comparison by NHS integrated care board")
```

# Maps

Darker, more opaque areas have a larger absolute percentage difference. Fill colour shows whether registered-patient counts are above or below ONS.

```{r region-map, fig.height=7}
region_map <- nhser_boundaries %>%
  left_join(region_snapshot, by = c("Boundary_Name" = "NHSER26NM"))

ggplot(region_map) +
  geom_polygon(
    aes(
      x = Longitude,
      y = Latitude,
      group = Polygon_Group,
      fill = Pct_Difference,
      alpha = abs(Pct_Difference)
    ),
    colour = "white",
    linewidth = 0.4
  ) +
  scale_fill_gradient2(
    low = "#B2182B",
    mid = "#F7F7F7",
    high = "#2166AC",
    midpoint = 0,
    labels = function(x) paste0(x, "%"),
    na.value = "grey90"
  ) +
  scale_alpha_continuous(range = c(0.3, 1), guide = "none") +
  labs(
    title = "Registered-patient difference from ONS by NHS England region",
    subtitle = format(snapshot_date, "%B %Y"),
    fill = "Difference"
  ) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "right")
```

```{r icb-map, fig.height=9}
icb_map <- icb_boundaries %>%
  left_join(icb_snapshot, by = c("Boundary_Code" = "ICB26CD"))

ggplot(icb_map) +
  geom_polygon(
    aes(
      x = Longitude,
      y = Latitude,
      group = Polygon_Group,
      fill = Pct_Difference,
      alpha = abs(Pct_Difference)
    ),
    colour = "white",
    linewidth = 0.15
  ) +
  scale_fill_gradient2(
    low = "#B2182B",
    mid = "#F7F7F7",
    high = "#2166AC",
    midpoint = 0,
    labels = function(x) paste0(x, "%"),
    na.value = "grey90"
  ) +
  scale_alpha_continuous(range = c(0.25, 1), guide = "none") +
  labs(
    title = "Registered-patient difference from ONS by NHS integrated care board",
    subtitle = format(snapshot_date, "%B %Y"),
    fill = "Difference"
  ) +
  coord_equal() +
  theme_void() +
  theme(legend.position = "right")
```

# Change in the raw population difference

The following charts show the year-on-year percentage change in the raw registered-minus-ONS difference between July 2023 and July 2024.

```{r region-raw-difference-change, fig.height=6}
ggplot(
  region_raw_change,
  aes(
    x = Raw_Difference_Pct_Change,
    y = reorder(NHSER26NM, Raw_Difference_Pct_Change),
    fill = Raw_Difference_Pct_Change
  )
) +
  geom_vline(xintercept = 0, colour = "grey55") +
  geom_col() +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "grey90", high = "#2166AC",
    midpoint = 0, guide = "none"
  ) +
  labs(
    title = "Change in the raw population difference by NHS region",
    subtitle = "July 2023 to July 2024",
    x = "Percentage change in registered minus ONS difference",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

```{r icb-raw-difference-change, fig.height=12}
ggplot(
  icb_raw_change,
  aes(
    x = Raw_Difference_Pct_Change,
    y = reorder(ICB26NM, Raw_Difference_Pct_Change),
    fill = Raw_Difference_Pct_Change
  )
) +
  geom_vline(xintercept = 0, colour = "grey55") +
  geom_col() +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "grey90", high = "#2166AC",
    midpoint = 0, guide = "none"
  ) +
  labs(
    title = "Change in the raw population difference by NHS ICB",
    subtitle = "July 2023 to July 2024",
    x = "Percentage change in registered minus ONS difference",
    y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_blank())
```

# Patients per qualified GP FTE

```{r patients-per-gp-prepare, include=FALSE}
workforce_national_july <- qualified_gp_workforce %>%
  filter(format(Period, "%m") == "07") %>%
  mutate(Year = as.integer(format(Period, "%Y"))) %>%
  group_by(Year) %>%
  summarise(GP_FTE = sum(GP_FTE, na.rm = TRUE), .groups = "drop")

population_national_july <- national_time_series %>%
  filter(format(Period, "%m") == "07") %>%
  mutate(Year = as.integer(format(Period, "%Y"))) %>%
  select(Year, Source, Population)

patients_per_gp_july <- population_national_july %>%
  inner_join(workforce_national_july, by = "Year") %>%
  mutate(
    Period = as.Date(paste0(Year, "-07-01")),
    Patients_Per_GP_FTE = Population / GP_FTE
  )

workforce_region_snapshot <- qualified_gp_workforce %>%
  filter(
    format(Period, "%Y-%m") == format(snapshot_date, "%Y-%m")
  ) %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(GP_FTE = sum(GP_FTE, na.rm = TRUE), .groups = "drop")

practice_region_registered <- practice_registered_snapshot %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(Registered = sum(Registered, na.rm = TRUE), .groups = "drop")

patients_per_gp_region_snapshot <- region_snapshot %>%
  transmute(NHS_Region = NHSER26NM, ONS) %>%
  full_join(practice_region_registered, by = "NHS_Region") %>%
  inner_join(workforce_region_snapshot, by = "NHS_Region") %>%
  mutate(
    Registered_Per_GP_FTE = Registered / GP_FTE,
    ONS_Per_GP_FTE = ONS / GP_FTE,
    Raw_Difference_Per_GP_FTE = Registered_Per_GP_FTE - ONS_Per_GP_FTE
  )

patients_per_gp_national_snapshot <- patients_per_gp_july %>%
  filter(Year == as.integer(format(snapshot_date, "%Y"))) %>%
  select(Source, Patients_Per_GP_FTE) %>%
  pivot_wider(names_from = Source, values_from = Patients_Per_GP_FTE) %>%
  transmute(
    Geography = "England",
    Registered_Per_GP_FTE = `Registered patients`,
    ONS_Per_GP_FTE = `ONS population estimate`,
    Raw_Difference_Per_GP_FTE =
      Registered_Per_GP_FTE - ONS_Per_GP_FTE
  )

patients_per_gp_difference <- bind_rows(
  patients_per_gp_national_snapshot %>%
    select(Geography, Raw_Difference_Per_GP_FTE),
  patients_per_gp_region_snapshot %>%
    transmute(
      Geography = NHS_Region,
      Raw_Difference_Per_GP_FTE
    )
)
```

```{r patients-per-gp-july-time-series}
ggplot(
  patients_per_gp_july,
  aes(x = Period, y = Patients_Per_GP_FTE, colour = Source)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    limits = c(0, NA),
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Patients per qualified GP FTE",
    subtitle = "England, July snapshots",
    x = NULL,
    y = "Patients per GP FTE",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )
```

```{r patients-per-gp-raw-difference, fig.height=7}
ggplot(
  patients_per_gp_difference,
  aes(
    x = Raw_Difference_Per_GP_FTE,
    y = reorder(Geography, Raw_Difference_Per_GP_FTE),
    fill = Raw_Difference_Per_GP_FTE
  )
) +
  geom_vline(xintercept = 0, colour = "grey55") +
  geom_col() +
  geom_text(
    aes(label = scales::comma(Raw_Difference_Per_GP_FTE, accuracy = 1)),
    hjust = ifelse(
      patients_per_gp_difference$Raw_Difference_Per_GP_FTE >= 0,
      -0.12,
      1.12
    )
  ) +
  scale_x_continuous(
    labels = scales::label_comma(),
    expand = expansion(mult = c(0.15, 0.18))
  ) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "grey90", high = "#2166AC",
    midpoint = 0, guide = "none"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Raw difference in patients per qualified GP FTE",
    subtitle = paste0(
      format(snapshot_date, "%B %Y"),
      "; registered-patient rate minus ONS rate"
    ),
    x = "Additional registered patients per GP FTE",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.margin = margin(5.5, 30, 5.5, 5.5)
  )
```

```{r patients-per-gp-table}
patients_per_gp_national_snapshot %>%
  transmute(
    Geography,
    `Registered patients per GP FTE` =
      scales::comma(Registered_Per_GP_FTE, accuracy = 1),
    `ONS population per GP FTE` =
      scales::comma(ONS_Per_GP_FTE, accuracy = 1),
    `Raw difference` =
      scales::comma(Raw_Difference_Per_GP_FTE, accuracy = 1)
  ) %>%
  knitr::kable(caption = "National patients per GP FTE comparison")

patients_per_gp_region_snapshot %>%
  arrange(desc(Raw_Difference_Per_GP_FTE)) %>%
  transmute(
    `NHS England region` = NHS_Region,
    `Registered patients per GP FTE` =
      scales::comma(Registered_Per_GP_FTE, accuracy = 1),
    `ONS population per GP FTE` =
      scales::comma(ONS_Per_GP_FTE, accuracy = 1),
    `Raw difference` =
      scales::comma(Raw_Difference_Per_GP_FTE, accuracy = 1)
  ) %>%
  knitr::kable(caption = "Regional patients per GP FTE comparison")
```

# Potential payment reduction

The payment scenario uses the latest available records for the metric Total NHS Payments to General Practice Minus Deductions from the payments query in 03_Payments_To_GP_Practice.R. It compares the observed payment per registered patient with the payment per ONS-estimated resident and estimates the reduction if the registered-patient payment rate were applied to the ONS population.

```{r payment-scenario-prepare, include=FALSE}
payment_metric_name <- "Total NHS Payments to General Practice Minus Deductions"

payment_data <- finance_datasets %>%
  mutate(
    Effective_Snapshot_Date = as.Date(Effective_Snapshot_Date),
    Metric_Value = as.numeric(Metric_Value)
  ) %>%
  filter(Metric == payment_metric_name)

payment_snapshot_date <- max(
  payment_data$Effective_Snapshot_Date,
  na.rm = TRUE
)

payment_by_practice <- payment_data %>%
  filter(Effective_Snapshot_Date == payment_snapshot_date) %>%
  group_by(Practice_Code) %>%
  summarise(
    Total_Payments = sum(Metric_Value, na.rm = TRUE),
    .groups = "drop"
  )

national_payment_scenario <- tibble::tibble(
  Geography = "England",
  Total_Payments = sum(payment_by_practice$Total_Payments, na.rm = TRUE),
  Registered = national_snapshot$Registered,
  ONS = national_snapshot$ONS
) %>%
  mutate(
    Payment_Per_Registered = Total_Payments / Registered,
    Payment_Per_ONS = Total_Payments / ONS,
    ONS_Adjusted_Payments = Payment_Per_Registered * ONS,
    Potential_Reduction = Total_Payments - ONS_Adjusted_Payments
  )

regional_payment_scenario <- payment_by_practice %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(
    Total_Payments = sum(Total_Payments, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(practice_region_registered, by = "NHS_Region") %>%
  left_join(
    region_snapshot %>% transmute(NHS_Region = NHSER26NM, ONS),
    by = "NHS_Region"
  ) %>%
  filter(Registered > 0, ONS > 0) %>%
  mutate(
    Payment_Per_Registered = Total_Payments / Registered,
    Payment_Per_ONS = Total_Payments / ONS,
    ONS_Adjusted_Payments = Payment_Per_Registered * ONS,
    Potential_Reduction = Total_Payments - ONS_Adjusted_Payments
  )
```

```{r national-payment-chart}
national_payment_long <- national_payment_scenario %>%
  select(
    Geography,
    `Registered-patient denominator` = Payment_Per_Registered,
    `ONS population denominator` = Payment_Per_ONS
  ) %>%
  pivot_longer(
    -Geography,
    names_to = "Denominator",
    values_to = "Payment_Per_Person"
  )

ggplot(
  national_payment_long,
  aes(x = Denominator, y = Payment_Per_Person, fill = Denominator)
) +
  geom_col(width = 0.65) +
  geom_text(
    aes(label = scales::label_currency(prefix = "£")(Payment_Per_Person)),
    vjust = -0.3
  ) +
  scale_y_continuous(
    labels = scales::label_currency(prefix = "£"),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.12))
  ) +
  scale_fill_manual(values = c("#2166AC", "#B2182B"), guide = "none") +
  labs(
    title = "National payment per person under alternative denominators",
    subtitle = paste(
      "Payment data snapshot:",
      format(payment_snapshot_date, "%B %Y"),
      "| Population snapshot:",
      format(snapshot_date, "%B %Y")
    ),
    x = NULL,
    y = "Payment per person"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())
```

```{r national-payment-table}
national_payment_scenario %>%
  transmute(
    Geography,
    `Total payments` =
      scales::dollar(Total_Payments, prefix = "£", accuracy = 1),
    `Payment per registered patient` =
      scales::dollar(Payment_Per_Registered, prefix = "£", accuracy = 0.01),
    `Payment per ONS resident` =
      scales::dollar(Payment_Per_ONS, prefix = "£", accuracy = 0.01),
    `Potential payment reduction` =
      scales::dollar(Potential_Reduction, prefix = "£", accuracy = 1)
  ) %>%
  knitr::kable(caption = "National payment scenario")
```

```{r regional-payment-chart, fig.height=7}
ggplot(
  regional_payment_scenario,
  aes(
    x = Potential_Reduction,
    y = reorder(NHS_Region, Potential_Reduction),
    fill = Potential_Reduction
  )
) +
  geom_vline(xintercept = 0, colour = "grey55") +
  geom_col() +
  scale_x_continuous(
    labels = scales::label_currency(
      prefix = "£",
      scale = 1e-6,
      suffix = "m"
    )
  ) +
  scale_fill_gradient2(
    low = "#B2182B", mid = "grey90", high = "#2166AC",
    midpoint = 0, guide = "none"
  ) +
  labs(
    title = "Potential payment reduction by NHS England region",
    subtitle = "Registered-patient payment rate applied to the ONS population",
    x = "Potential reduction",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

```{r regional-payment-table}
regional_payment_scenario %>%
  arrange(desc(Potential_Reduction)) %>%
  transmute(
    `NHS England region` = NHS_Region,
    `Total payments` =
      scales::dollar(Total_Payments, prefix = "£", accuracy = 1),
    `Payment per registered patient` =
      scales::dollar(Payment_Per_Registered, prefix = "£", accuracy = 0.01),
    `Payment per ONS resident` =
      scales::dollar(Payment_Per_ONS, prefix = "£", accuracy = 0.01),
    `Potential payment reduction` =
      scales::dollar(Potential_Reduction, prefix = "£", accuracy = 1)
  ) %>%
  knitr::kable(caption = "Regional payment scenario")
```

## Data sources

- GP registered-patient national, age, and sex data: Demography.No_Of_Patients_Regd_At_GP_Practice_Single_Age1.
- GP registered-patient LSOA data: Demography.No_Of_Patients_Regd_At_GP_Practice_LSOA_2021_Level1.
- ONS population estimates: Demography.ONS_Population_Estimates_For_LSOAs_By_Year_Of_Age1.
- LSOA IMD: latest snapshot in Demography.Index_Of_Multiple_Deprivation_By_LSOA1.
- GP workforce: NHS_Workforce.Practice_Level_Census_Data1, using TOTAL_GP_EXTG_FTE.
- GP payments: NHS_Payments_To_GP_Practices.Figures1 via 03_Payments_To_GP_Practice.R.
- LSOA-to-ICB and NHS-region lookup: ONS April 2026 lookup.
- Practice-to-region lookup: payments2425.csv.
- Boundary maps: ONS Open Geography Portal generalised boundaries.
