---
title: "Ghost Patients Descriptive Analysis"
output:
  word_document:
    reference_docx: "05_Patient_List_Focused_Clean.docx"
  html_document:
    toc: true
    toc_float: true
params:
  start_date: "2015-07-01"
  end_date: "2026-08-31"
  snapshot_date: "2024-07-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5.5,
  dpi = 150
)

required_packages <- c(
  "DBI", "odbc", "dplyr", "tidyr", "ggplot2", "scales",
  "knitr", "httr", "jsonlite", "readr", "glue", "rlang", "tibble",
  "tidyverse"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the following R packages before rendering: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(DBI)
library(odbc)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(knitr)

start_date <- as.Date(params$start_date)
end_date <- as.Date(params$end_date)
snapshot_date <- as.Date(params$snapshot_date)

stopifnot(
  !is.na(start_date),
  !is.na(end_date),
  !is.na(snapshot_date),
  start_date <= end_date
)

pct_difference <- function(registered, ons) {
  ifelse(
    is.na(ons) | ons == 0,
    NA_real_,
    100 * (registered - ons) / ons
  )
}

# Build the pound sign at render time to avoid escaped Unicode text in charts.
gbp_symbol <- intToUtf8(163)

comparison_bar <- function(
  data,
  category,
  title,
  subtitle,
  y_title = NULL,
  fixed_fill = NULL
) {
  category_quo <- rlang::enquo(category)

  p <- ggplot(
    data,
    aes(
      x = Pct_Difference,
      y = !!category_quo
    )
  ) +
    geom_vline(xintercept = 0, colour = "grey55", linewidth = 0.5)

  if (is.null(fixed_fill)) {
    p <- p +
      geom_col(aes(fill = Pct_Difference), width = 0.7) +
      scale_fill_gradient2(
        low = "#B2182B",
        mid = "#F2F2F2",
        high = "#2166AC",
        midpoint = 0,
        guide = "none"
      )
  } else {
    p <- p + geom_col(fill = fixed_fill, width = 0.7)
  }

  p +
    geom_text(
    colour = "grey20",
      aes(label = sprintf("%.1f%%", Pct_Difference)),
      hjust = ifelse(data$Pct_Difference >= 0, -0.12, 1.12),
      size = 3.7
    ) +
    scale_x_continuous(
      labels = function(x) paste0(x, "%"),
      expand = expansion(mult = c(0.16, 0.18))
    ) +
    coord_cartesian(clip = "off") +
    labs(
      title = title,
      subtitle = subtitle,
      x = "Percentage difference from ONS",
      y = y_title,
      caption = "Percentage difference = (registered patients − ONS estimate) / ONS estimate × 100."
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = margin(5.5, 30, 5.5, 5.5)
    )
}


# HTML tables start closed; document outputs retain ordinary visible tables.
report_table <- function(x, caption = NULL, ...) {
  if (!knitr::is_html_output()) {
    return(knitr::kable(x, caption = caption, ...))
  }
  summary <- if (is.null(caption)) "View table" else paste("View table:", caption)
  summary <- gsub("&", "&amp;", summary, fixed = TRUE)
  summary <- gsub("<", "&lt;", summary, fixed = TRUE)
  summary <- gsub(">", "&gt;", summary, fixed = TRUE)
  table_html <- paste(
    knitr::kable(x, format = "html", caption = caption, ...),
    collapse = "\n"
  )
  knitr::asis_output(paste0(
    "\n<details class=\"report-table\"><summary>", summary, "</summary>\n",
    "<div class=\"report-table-scroll\">\n", table_html,
    "\n</div>\n</details>\n"
  ))
}

# Source labels are embedded in figures, so exported charts retain provenance.
label_chart <- function(plot, source, inputs = NULL) {
  source_text <- paste0("Registration data: ", source,
                        ". ONS: English LSOA population estimates.")
  if (!is.null(inputs)) source_text <- paste(source_text, inputs)
  existing_caption <- plot$labels$caption
  caption_lines <- vapply(c(source_text, existing_caption), function(x) {
    paste(strwrap(x, width = 105), collapse = "\n")
  }, character(1))
  plot + labs(caption = paste(caption_lines, collapse = "\n")) +
    theme(plot.caption = element_text(colour = "grey20", hjust = 0, size = 9))
}

first_matching_column <- function(data, exact = character(), regex = NULL) {
  nm <- names(data)

  exact_match <- exact[exact %in% nm]
  if (length(exact_match) > 0) {
    return(exact_match[[1]])
  }

  if (!is.null(regex)) {
    regex_match <- nm[grepl(regex, nm, ignore.case = TRUE)]
    if (length(regex_match) > 0) {
      return(regex_match[[1]])
    }
  }

  NA_character_
}
```

```{r report-table-style, echo=FALSE, results='asis'}
if (knitr::is_html_output()) {
  cat('<style>
.report-table { margin: 1em 0; border: 1px solid #d8dfe5; border-radius: 4px; padding: .6em .8em; }
.report-table > summary { cursor: pointer; font-weight: 600; color: #2166ac; }
.report-table > summary:focus-visible { outline: 2px solid #2166ac; outline-offset: 3px; }
.report-table-scroll { overflow-x: auto; margin-top: .8em; }
.report-table-scroll table { width: 100%; }
</style>')
}
```

```{r connect-and-load-data, include=FALSE}
source("01_connect_to_facts_and_dimensions.R")
source("03_Payments_To_GP_Practice.R")
source("08_Get_Focused_Patient_List_Data.R")
source("09_Get_LSOA_NHS_Region_Lookup.R")
source("10_Get_Practice_Region_Lookup.R")
DBI::dbDisconnect(con)
source("12_Summarise_Registration_Coverage.R")
source("14_Summarise_IMD_Trends.R")
```

```{r prepare-analysis-data, include=FALSE}
age_levels <- c(
  "0-9", "10-19", "20-29", "30-39", "40-49",
  "50-59", "60-69", "70-79", "80-89", "90+"
)

national_time_series <- bind_rows(national_registered, national_ons)

demographic_time_series <- full_join(
  practice_registered_demographic_time_series,
  ons_demographic_time_series,
  by = c("Period", "Sex", "Age_Band")
) %>%
  mutate(
    Age_Band = factor(Age_Band, levels = age_levels),
    Sex = factor(Sex, levels = c("Female", "Male")),
    Pct_Difference = pct_difference(Registered, ONS)
  )

sum_if_available <- function(x) {
  if (all(is.na(x))) NA_real_ else sum(x, na.rm = TRUE)
}

age_time_series <- demographic_time_series %>%
  group_by(Period, Age_Band) %>%
  summarise(
    Registered = sum_if_available(Registered),
    ONS = sum_if_available(ONS),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Registered, ONS),
    names_to = "Source",
    values_to = "Population"
  ) %>%
  mutate(
    Source = recode(
      Source,
      Registered = "Registered patients",
      ONS = "ONS population estimate"
    )
  )

ons_sex_time_series <- ons_demographic_time_series %>%
  group_by(Period, Sex) %>%
  summarise(ONS = sum_if_available(ONS), .groups = "drop")

sex_comparison <- full_join(
  practice_registered_demographic_time_series %>%
    group_by(Period, Sex) %>%
    summarise(Registered = sum_if_available(Registered), .groups = "drop"),
  ons_sex_time_series, by = c("Period", "Sex")
) %>% mutate(Pct_Difference = pct_difference(Registered, ONS))

sex_time_series <- sex_comparison %>%
  pivot_longer(
    cols = c(Registered, ONS),
    names_to = "Source",
    values_to = "Population"
  ) %>%
  mutate(
    Source = recode(
      Source,
      Registered = "Registered patients",
      ONS = "ONS population estimate"
    )
  )

national_snapshot <- national_time_series %>%
  filter(Period == snapshot_date) %>%
  select(Source, Population) %>%
  pivot_wider(names_from = Source, values_from = Population) %>%
  transmute(
    Geography = "England",
    Registered = `Registered patients`,
    ONS = `ONS population estimate`,
    Pct_Difference = pct_difference(Registered, ONS)
  )

if (nrow(national_snapshot) != 1L ||
    any(!is.finite(c(national_snapshot$Registered, national_snapshot$ONS)))) {
  stop("Both practice registrations and English ONS estimates are required at the selected snapshot.")
}

age_sex_snapshot <- demographic_time_series %>%
  filter(Period == snapshot_date)

age_snapshot <- age_sex_snapshot %>%
  group_by(Age_Band) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Pct_Difference = pct_difference(Registered, ONS))

sex_snapshot <- sex_comparison %>%
  filter(Period == snapshot_date) %>%
  group_by(Sex) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Pct_Difference = pct_difference(Registered, ONS))

lsoa_july_all <- full_join(
  lsoa_registered_july,
  lsoa_ons_july,
  by = c("LSOA_Code", "Period")
) %>%
  mutate(
    Registered = replace_na(Registered, 0),
    ONS = replace_na(ONS, 0),
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

# Explicit England-resident subset for every LSOA-derived analysis.
lsoa_july_england <- lsoa_july_all %>%
  filter(grepl("^E01", LSOA_Code))

# Quantify registered patients excluded from the England-resident LSOA subset.
non_england_registered_snapshot <- lsoa_registered_july %>%
  filter(
    Period == snapshot_date,
    !grepl("^E01", LSOA_Code)
  ) %>%
  mutate(
    Residence_Group = case_when(
      grepl("^W0", LSOA_Code) ~ "Wales",
      TRUE ~ "Other non-English LSOA"
    )
  ) %>%
  group_by(Residence_Group) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    .groups = "drop"
  )

registered_removed_wales <- non_england_registered_snapshot %>%
  filter(Residence_Group == "Wales") %>%
  summarise(Registered = sum(Registered, na.rm = TRUE)) %>%
  pull(Registered)

registered_removed_other <- non_england_registered_snapshot %>%
  filter(Residence_Group == "Other non-English LSOA") %>%
  summarise(Registered = sum(Registered, na.rm = TRUE)) %>%
  pull(Registered)

registered_removed_non_england <-
  registered_removed_wales + registered_removed_other

lsoa_registered_snapshot_total <- lsoa_registered_july %>%
  filter(Period == snapshot_date) %>%
  summarise(Registered = sum(Registered, na.rm = TRUE)) %>%
  pull(Registered)

registered_removed_non_england_pct <-
  100 * registered_removed_non_england / lsoa_registered_snapshot_total

if (nrow(anti_join(
  lsoa_july_england, nhser_lookup, by = c("LSOA_Code" = "LSOA21CD")
)) > 0) {
  stop("Some English LSOAs are missing from the April 2026 geography lookup.")
}

lsoa_july_geography <- lsoa_july_england %>%
  inner_join(nhser_lookup, by = c("LSOA_Code" = "LSOA21CD"))

registration_coverage <- summarise_registration_coverage(
  national_registered, lsoa_registered_july, imd_lookup, snapshot_date
)

imd_july <- lsoa_july_england %>%
  inner_join(
    imd_lookup %>% mutate(
      IMD_Decile = suppressWarnings(as.numeric(as.character(IMD_Decile)))
    ),
    by = "LSOA_Code"
  ) %>%
  filter(IMD_Decile %in% 1:10) %>%
  group_by(Period, IMD_Decile) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS),
    IMD_Label = factor(
      IMD_Decile,
      levels = 10:1,
      labels = c(
        "10 (Least deprived)", "9", "8", "7", "6",
        "5", "4", "3", "2", "1 (Most deprived)"
      )
    )
  )

region_july <- lsoa_july_geography %>%
  group_by(Period, NHSER26CD, NHSER26NM) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

icb_july <- lsoa_july_geography %>%
  group_by(Period, ICB26CD, ICB26NM) %>%
  summarise(
    Registered = sum(Registered, na.rm = TRUE),
    ONS = sum(ONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Raw_Difference = Registered - ONS,
    Pct_Difference = pct_difference(Registered, ONS)
  )

imd_snapshot <- imd_july %>% filter(Period == snapshot_date)
if (abs(sum(imd_snapshot$Registered) - registration_coverage$imd_patients) > 0.5) {
  stop("The deprivation analysis does not reconcile with the opening coverage summary.")
}
region_snapshot <- region_july %>% filter(Period == snapshot_date)
icb_snapshot <- icb_july %>% filter(Period == snapshot_date)

required_practice_lookup_cols <- c("Practice_Code", "NHS_Region")
missing_region_lookup_cols <- setdiff(
  required_practice_lookup_cols,
  names(practice_region_lookup)
)

if (length(missing_region_lookup_cols) > 0) {
  stop(
    "practice_region_lookup is missing required columns: ",
    paste(missing_region_lookup_cols, collapse = ", ")
  )
}

required_icb_lookup_cols <- c("Practice_Code", "ICB_Key", "ICB")
missing_icb_lookup_cols <- setdiff(
  required_icb_lookup_cols,
  names(practice_icb_lookup)
)

if (length(missing_icb_lookup_cols) > 0) {
  stop(
    "practice_icb_lookup is missing required columns: ",
    paste(missing_icb_lookup_cols, collapse = ", ")
  )
}

if (nrow(practice_icb_lookup) == 0) {
  stop(
    "The practice-to-ICB lookup is empty. Check the practice ICB fields ",
    "in payments2425.csv and 10_Get_Practice_Region_Lookup.R."
  )
}
# Registrations follow practice assignment; ONS continues to follow residence.
# A practice must map to at most one geography to prevent duplicate counting.
if (anyDuplicated(practice_region_lookup$Practice_Code) ||
    anyDuplicated(practice_icb_lookup$Practice_Code)) {
  stop("Practice geography lookups contain duplicate practice codes.")
}
practice_region_registered <- practice_registered_snapshot %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(Registered = sum(Registered), .groups = "drop")

practice_icb_registered <- practice_registered_snapshot %>%
  inner_join(practice_icb_lookup, by = "Practice_Code") %>%
  group_by(ICB_Key, ICB) %>%
  summarise(Registered = sum(Registered), .groups = "drop")

# Population comparison charts retain residence-based LSOA registrations.
region_ons_keys <- region_snapshot %>%
  mutate(NHS_Region = normalise_region_name(NHSER26NM))
if (nrow(anti_join(practice_region_registered, region_ons_keys, by = "NHS_Region")) ||
    nrow(anti_join(practice_icb_registered, icb_snapshot,
                   by = c("ICB_Key" = "ICB26CD")))) {
  stop("A practice-assigned geography has no matching ONS population geography.")
}

registered_unassigned_region <- national_snapshot$Registered -
  sum(practice_region_registered$Registered)
registered_unassigned_icb <- national_snapshot$Registered -
  sum(practice_icb_registered$Registered)
for (local in list(region_snapshot, icb_snapshot)) {
  if (any(!is.finite(local$Registered)) ||
      abs(sum(local$Registered) - sum(lsoa_july_england$Registered)) > 0.5 ||
      abs(sum(local$ONS) - national_snapshot$ONS) > 0.5) {
    stop("Geographic LSOA comparison totals do not reconcile with their sources.")
  }
}
if (registered_unassigned_region < -0.5 || registered_unassigned_icb < -0.5) {
  stop("Assigned practice registrations exceed the national practice source.")
}
```

# Dataset Disparities

For `r format(snapshot_date, "%d %B %Y")` data, the practice-level registered patients data contains `r scales::comma(registration_coverage$practice_total)` registered patients, compared with `r scales::comma(registration_coverage$lsoa_total)` in the LSOA-level registered patients’ data before any filtering or linkage. The LSOA source contains `r scales::comma(abs(registration_coverage$practice_total - registration_coverage$lsoa_total))` `r ifelse(registration_coverage$lsoa_total <= registration_coverage$practice_total, "fewer", "more")` registrations than the practice-level total. [NHS England reported processing issues in the July 2024 LSOA release](https://digital.nhs.uk/data-and-information/publications/statistical/patients-registered-at-a-gp-practice/july-2024) which led to this difference.

The LSOA extract also includes `r scales::comma(registration_coverage$welsh_patients)` registered patients in `r scales::comma(registration_coverage$welsh_lsoa_count)` Welsh LSOAs. Another `r scales::comma(registration_coverage$other_patients)` registrations have an unknown or other residence code and are also excluded from the LSOA comparison dataset, leaving `r scales::comma(registration_coverage$english_patients)` registrations in the English LSOAs dataset used for IMD, regional and ICB population comparisons. Welsh LSOAs have no English IMD. English registrations without a valid IMD decile exclude another `r scales::comma(registration_coverage$english_patients_without_imd)` patients from the IMD comparison.

IMD, regional and ICB population comparisons use residence-based LSOA registrations. The new deprivation trends also use LSOA data. All other registration charts and tables, including FTE and funding analyses, use practice-level registrations. Practice registrations include patients of English GP practices irrespective of residence; Welsh and unknown LSOA residence codes are therefore retained in the practice-based totals. The residence exclusions above apply to all LSOA population comparisons; the additional missing-IMD exclusion applies only to deprivation analyses. ONS estimates cover English-resident LSOAs. In the FTE and funding sections, regional and April 2026 ICB registrations, workforce and payments follow practice assignment, while ONS populations follow residence. Aggregate registration differences do not identify individual ghost patients.

# National population over time

```{r national-population-time-series}
label_chart(
ggplot(
  national_time_series,
  aes(x = Period, y = Population, colour = Source)
) +
  geom_line(linewidth = 1.1) +
  geom_point(
    data = national_time_series %>%
      filter(Source == "ONS population estimate"),
    size = 2
  ) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(
    limits = c(start_date, end_date),
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "m")
  ) +
  labs(
    title = "England population: registered patients and ONS estimates",
    subtitle = "July 2015 onwards; available source snapshots",
    x = NULL,
    y = "Population",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ),
  "Practice-level registrations"
)
```

# Age over time

```{r age-time-series, fig.height=10}
label_chart(
ggplot(
  age_time_series,
  aes(x = Period, y = Population, colour = Source, group = Source)
) +
  geom_vline(
    xintercept = as.numeric(as.Date("2021-07-01")),
    linetype = "dashed", colour = "grey65", linewidth = 0.4
  ) +
  geom_line(
    data = age_time_series %>% filter(!is.na(Population)),
    linewidth = 0.8
  ) +
  geom_point(
    data = age_time_series %>%
      filter(Source == "ONS population estimate", !is.na(Population)),
    size = 1.3
  ) +
  facet_wrap(~ Age_Band, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "m")) +
  labs(
    title = "Population by age over time",
    subtitle = "English GP registrations and England-only ONS estimates; matching ten-year age bands",
    x = NULL,
    y = "Population",
    colour = NULL,
    caption = "Dashed line: Census 2021 year; the cause of any turning point is not established by the chart."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ),
  "Practice-level registrations"
)
```

# Sex over time

```{r sex-time-series, fig.height=7}
label_chart(
ggplot(
  sex_time_series,
  aes(x = Period, y = Population, colour = Source, group = Source)
) +
  geom_line(
    data = sex_time_series %>% filter(!is.na(Population)),
    linewidth = 1
  ) +
  geom_point(
    data = sex_time_series %>%
      filter(Source == "ONS population estimate", !is.na(Population)),
    size = 1.8
  ) +
  facet_wrap(~ Sex, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "m")) +
  labs(
    title = "Population by sex over time",
    subtitle = "Registered patients and ONS estimates",
    x = NULL,
    y = "Population",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ),
  "Practice-level registrations"
)
```

# July 2024 Snapshots

## National total

```{r national-snapshot-chart}
label_chart(
comparison_bar(
  national_snapshot,
  Geography,
  "National percentage difference",
  format(snapshot_date, "%B %Y"),
  fixed_fill = "#2166AC"
),
  "Practice-level registrations"
)
```

```{r national-snapshot-table, results='asis'}
national_snapshot %>%
  transmute(
    Geography,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "National population comparison (Practice-level registrations)")
```

## Age

```{r age-snapshot-chart, fig.height=6}
label_chart(
comparison_bar(
  age_snapshot,
  Age_Band,
  "Percentage difference by age",
  format(snapshot_date, "%B %Y"),
  "Age band"
),
  "Practice-level registrations"
)
```

```{r age-snapshot-table, results='asis'}
age_snapshot %>%
  transmute(
    `Age band` = Age_Band,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by age (Practice-level registrations)")
```

## Sex

```{r sex-snapshot-chart}
label_chart(
comparison_bar(
  sex_snapshot,
  Sex,
  "Percentage difference by sex",
  format(snapshot_date, "%B %Y"),
  "Sex"
),
  "Practice-level registrations"
)
```

```{r sex-snapshot-table, results='asis'}
sex_snapshot %>%
  transmute(
    Sex,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by sex (Practice-level registrations)")
```

## Age and sex

```{r age-sex-snapshot-chart, fig.height=7}
label_chart(
ggplot(
  age_sex_snapshot,
  aes(x = Age_Band, y = Pct_Difference, fill = Sex)
) +
  geom_hline(yintercept = 0, colour = "grey55") +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(
    colour = "grey20",
    aes(
      label = sprintf("%.1f%%", Pct_Difference),
      group = Sex,
      vjust = if_else(Pct_Difference >= 0, -0.35, 1.35)
    ),
    position = position_dodge(width = 0.8),
    size = 3.4
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0.12, 0.14))
  ) +
  coord_cartesian(clip = "off") +
  scale_fill_manual(values = c(Female = "#CC79A7", Male = "#0072B2")) +
  labs(
    title = "Percentage difference by age and sex",
    subtitle = format(snapshot_date, "%B %Y"),
    x = "Age band",
    y = "Percentage difference from ONS",
    fill = "Sex"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ),
  "Practice-level registrations"
)
```

```{r age-sex-snapshot-table, results='asis'}
age_sex_snapshot %>%
  arrange(Age_Band, Sex) %>%
  transmute(
    `Age band` = Age_Band,
    Sex,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Registered - ONS),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by age and sex (Practice-level registrations)")
```

# IMD deciles using LSOA data

```{r imd-snapshot-chart, fig.height=6}
label_chart(
comparison_bar(
  imd_snapshot,
  IMD_Label,
  "Percentage difference by LSOA-level IMD decile",
  format(snapshot_date, "%B %Y"),
  "IMD decile"
),
  "LSOA-level registrations"
)
```

This excludes: `r scales::comma(registered_removed_wales)` in Wales and `r scales::comma(registered_removed_other)` in “other LSOAs”. It also excludes `r scales::comma(registration_coverage$english_patients_without_imd)` English registrations without a valid IMD decile.

```{r imd-snapshot-table, results='asis'}
imd_snapshot %>%
  arrange(IMD_Decile) %>%
  transmute(
    `IMD decile` = IMD_Label,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by IMD decile (LSOA-level registrations)")
```

# Population over time by IMD quintile

These two charts use July snapshots and one fixed IMD lookup, pairing deciles 1–2 into quintile 1 (most deprived), 3–4 into quintile 2, through 9–10 into quintile 5 (least deprived). The classification is applied retrospectively; these are not historical IMD rankings.

```{r imd-quintile-trend-prepare, include=FALSE}
imd_trends <- summarise_imd_trends(
  bind_rows(lsoa_registered_history, lsoa_ons_history), imd_lookup
)
imd_trend_data <- imd_trends$trend %>%
  mutate(Quintile = factor(Quintile, levels = 1:5,
    labels = c("1 (Most deprived)", "2", "3", "4", "5 (Least deprived)"))) %>%
  group_by(Source, Quintile) %>%
  arrange(Period, .by_group = TRUE) %>%
  mutate(Year = as.integer(format(Period, "%Y")),
         Run = cumsum(c(TRUE, diff(Year) > 1))) %>%
  ungroup()
imd_trend_source_date <- paste(sort(unique(as.character(imd_lookup$IMD_Source_Date))),
                               collapse = ", ")
imd_trend_chart <- function(source_name, title) {
  chart_data <- imd_trend_data %>% filter(Source == source_name)
  ggplot(chart_data, aes(Period, Population, colour = Quintile)) +
    geom_line(aes(group = interaction(Quintile, Run)), linewidth = 0.9) +
    geom_point(size = 2) +
    scale_colour_manual(values = c("#7B3294", "#008837", "#D95F02", "#2166AC", "#A6761D"),
                        drop = FALSE) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    scale_y_continuous(labels = scales::label_number(scale = 1e-6, suffix = "m")) +
    labs(title = title,
      subtitle = paste("July snapshots:", format(min(chart_data$Period), "%Y"),
                       "to", format(max(chart_data$Period), "%Y")),
      x = NULL, y = "Population in common LSOA cohort", colour = "IMD quintile",
      caption = paste0(
        "Source: ", if (source_name == "Registered patients")
          "LSOA-level registrations by English residence." else
          "ONS English LSOA population estimates.",
        "\nFixed IMD classification; common LSOAs in every available snapshot of both sources.",
        "\nMissing years are not imputed or joined across.")) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1),
          plot.caption = element_text(colour = "grey20", hjust = 0))
}
```

The shared cohort contains **`r scales::comma(length(imd_trends$lsoa_codes))` English LSOAs** with valid IMD and data in every available source snapshot. These counts describe that cohort, not the full English population. Welsh/unknown residence codes, invalid IMD and LSOAs missing from any source snapshot are excluded. The warehouse IMD snapshot is `r imd_trend_source_date`; this date alone does not establish the index edition. The coverage table shows the population retained and excluded at each date.

```{r imd-quintile-ons-time-series, fig.height=6}
imd_trend_chart("ONS population estimate", "ONS resident population over time by IMD quintile")
```

```{r imd-quintile-registered-time-series, fig.height=6}
imd_trend_chart("Registered patients", "Registered patients over time by IMD quintile")
```

```{r imd-quintile-coverage-table, results='asis'}
imd_trends$coverage %>%
  transmute(Source, Snapshot = format(Period, "%d %b %Y"),
    `English n` = scales::comma(English_Population),
    `IMD-linked n` = scales::comma(IMD_Eligible_Population),
    `Cohort n` = scales::comma(Cohort_Population),
    `Excluded n` = scales::comma(Population_Excluded),
    `Cohort LSOAs` = scales::comma(Cohort_LSOAs)) %>%
  report_table(caption = "LSOA deprivation trend coverage")
```

# NHS England regions

```{r region-snapshot-chart, fig.height=6}
region_snapshot_plot <- region_snapshot %>%
  mutate(NHSER26NM = reorder(NHSER26NM, Pct_Difference))

label_chart(
comparison_bar(
  region_snapshot_plot,
  NHSER26NM,
  "Percentage difference by NHS England region",
  format(snapshot_date, "%B %Y"),
  "NHS England region"
),
  "LSOA-level registrations"
)
```

```{r region-snapshot-table, results='asis'}
region_snapshot %>%
  arrange(desc(Pct_Difference)) %>%
  transmute(
    `NHS England region` = NHSER26NM,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by NHS England region (LSOA-level registrations)")
```

# Integrated Care Boards

All ICB comparisons use the new April 2026 structure while using the July 2024 data. This population chart groups both registered patients and ONS residents through the official ONS April 2026 LSOA lookup, including the Frimley boundary changes. The later workforce and funding sections instead group practice registrations, workforce and payments through the practice-to-2026-ICB mapping.

```{r icb-snapshot-chart, fig.height=12}
icb_snapshot_plot <- icb_snapshot %>%
  mutate(ICB26NM = reorder(ICB26NM, Pct_Difference))

label_chart(
comparison_bar(
  icb_snapshot_plot,
  ICB26NM,
  "Percentage difference by NHS integrated care board",
  format(snapshot_date, "%B %Y"),
  "Integrated care board"
),
  "LSOA-level registrations"
)
```

```{r icb-snapshot-table, results='asis'}
icb_snapshot %>%
  arrange(desc(Pct_Difference)) %>%
  transmute(
    ICB = ICB26NM,
    `Registered patients` = scales::comma(Registered),
    `ONS population` = scales::comma(ONS),
    `Raw difference` = scales::comma(Raw_Difference),
    `Percentage difference` = sprintf("%.1f%%", Pct_Difference)
  ) %>%
  report_table(caption = "Population comparison by NHS integrated care board (LSOA-level registrations)")
```

# Patients per qualified GP FTE

```{r patients-per-gp-prepare, include=FALSE}
workforce_national_july <- qualified_gp_workforce %>%
  filter(format(Period, "%m") == "07") %>%
  mutate(Year = as.integer(format(Period, "%Y"))) %>%
  group_by(Year) %>%
  summarise(GP_FTE = sum(GP_FTE, na.rm = TRUE), .groups = "drop")

population_national_july <- national_time_series %>%
  filter(format(Period, "%m") == "07") %>%
  mutate(Year = as.integer(format(Period, "%Y"))) %>%
  select(Year, Source, Population)

patients_per_gp_july <- population_national_july %>%
  inner_join(workforce_national_july, by = "Year") %>%
  mutate(
    Period = as.Date(paste0(Year, "-07-01")),
    Patients_Per_GP_FTE = Population / GP_FTE
  )

workforce_snapshot <- qualified_gp_workforce %>%
  filter(format(Period, "%Y-%m") == format(snapshot_date, "%Y-%m"))

workforce_region_snapshot <- workforce_snapshot %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(GP_FTE = sum(GP_FTE, na.rm = TRUE), .groups = "drop")

patients_per_gp_region_snapshot <- region_snapshot %>%
  transmute(
    NHS_Region = normalise_region_name(NHSER26NM),
    NHS_Region_Display = NHSER26NM,
    ONS
  ) %>%
  full_join(practice_region_registered, by = "NHS_Region") %>%
  inner_join(workforce_region_snapshot, by = "NHS_Region") %>%
  filter(GP_FTE > 0) %>%
  mutate(
    Registered_Per_GP_FTE = Registered / GP_FTE,
    ONS_Per_GP_FTE = ONS / GP_FTE,
    Pct_Difference_Per_GP_FTE =
      100 * (Registered_Per_GP_FTE - ONS_Per_GP_FTE) / ONS_Per_GP_FTE
  )

workforce_icb_snapshot <- workforce_snapshot %>%
  inner_join(practice_icb_lookup, by = "Practice_Code") %>%
  group_by(ICB_Key, ICB) %>%
  summarise(GP_FTE = sum(GP_FTE, na.rm = TRUE), .groups = "drop")

icb_population_snapshot <- icb_snapshot %>%
  transmute(
    ICB_Key = ICB26CD,
    ICB_Population_Name = ICB26NM,
    ONS
  )

patients_per_gp_icb_snapshot <- icb_population_snapshot %>%
  full_join(
    practice_icb_registered %>% rename(ICB_Registered_Name = ICB),
    by = "ICB_Key"
  ) %>%
  inner_join(
    workforce_icb_snapshot %>% rename(ICB_Workforce_Name = ICB),
    by = "ICB_Key"
  ) %>%
  filter(GP_FTE > 0) %>%
  mutate(
    ICB = coalesce(
      ICB_Population_Name,
      ICB_Registered_Name,
      ICB_Workforce_Name
    ),
    Registered_Per_GP_FTE = Registered / GP_FTE,
    ONS_Per_GP_FTE = ONS / GP_FTE
  ) %>%
  select(
    ICB_Key,
    ICB,
    Registered,
    ONS,
    GP_FTE,
    Registered_Per_GP_FTE,
    ONS_Per_GP_FTE
  )

patients_per_gp_national_snapshot <- patients_per_gp_july %>%
  filter(Year == as.integer(format(snapshot_date, "%Y"))) %>%
  select(Source, Patients_Per_GP_FTE) %>%
  pivot_wider(names_from = Source, values_from = Patients_Per_GP_FTE) %>%
  transmute(
    Geography = "England",
    Registered_Per_GP_FTE = `Registered patients`,
    ONS_Per_GP_FTE = `ONS population estimate`
  )

patients_per_gp_region_long <- patients_per_gp_region_snapshot %>%
  select(
    NHS_Region,
    `Registered patients` = Registered_Per_GP_FTE,
    `ONS population` = ONS_Per_GP_FTE
  ) %>%
  pivot_longer(
    -NHS_Region,
    names_to = "Population basis",
    values_to = "Patients_Per_GP_FTE"
  )

region_gp_order <- patients_per_gp_region_snapshot %>%
  arrange(Registered_Per_GP_FTE) %>%
  pull(NHS_Region)

patients_per_gp_region_long <- patients_per_gp_region_long %>%
  mutate(NHS_Region = factor(NHS_Region, levels = region_gp_order))

patients_per_gp_icb_long <- patients_per_gp_icb_snapshot %>%
  select(
    ICB,
    `Registered patients` = Registered_Per_GP_FTE,
    `ONS population` = ONS_Per_GP_FTE
  ) %>%
  pivot_longer(
    -ICB,
    names_to = "Population basis",
    values_to = "Patients_Per_GP_FTE"
  )

icb_gp_order <- patients_per_gp_icb_snapshot %>%
  arrange(Registered_Per_GP_FTE) %>%
  pull(ICB)

patients_per_gp_icb_long <- patients_per_gp_icb_long %>%
  mutate(ICB = factor(ICB, levels = icb_gp_order))
```

Registrations and qualified GP FTE follow practice assignment to the April 2026 geography. ONS populations follow English residence LSOAs. Regional and ICB comparisons can therefore reflect cross-boundary registration. Unassigned practice registrations are excluded from the corresponding local comparisons: `r scales::comma(registered_unassigned_region)` without a region and `r scales::comma(registered_unassigned_icb)` without an ICB.

## July 2024 national comparison

```{r patients-per-gp-july-2024-bar, fig.width=8, fig.height=5}
patients_per_gp_july_2024 <- patients_per_gp_july %>%
  filter(Period == as.Date("2024-07-01"))

expected_population_sources <- c(
  "ONS population estimate", "Registered patients"
)

if (nrow(patients_per_gp_july_2024) != 2L ||
    !setequal(patients_per_gp_july_2024$Source, expected_population_sources) ||
    any(!is.finite(patients_per_gp_july_2024$GP_FTE)) ||
    any(patients_per_gp_july_2024$GP_FTE <= 0) ||
    n_distinct(patients_per_gp_july_2024$GP_FTE) != 1L ||
    any(!is.finite(patients_per_gp_july_2024$Patients_Per_GP_FTE)) ||
    any(patients_per_gp_july_2024$Patients_Per_GP_FTE < 0)) {
  stop("The July 2024 bar chart requires both population sources and one positive qualified GP FTE total.")
}

patients_per_gp_july_2024 <- patients_per_gp_july_2024 %>%
  mutate(Source = factor(Source, levels = expected_population_sources))

label_chart(
ggplot(
  patients_per_gp_july_2024,
  aes(x = Source, y = Patients_Per_GP_FTE, fill = Source)
) +
  geom_col(width = 0.6) +
  geom_text(
    colour = "grey20",
    aes(label = scales::comma(Patients_Per_GP_FTE, accuracy = 1)),
    vjust = -0.5,
    size = 4
  ) +
  scale_fill_manual(values = c(
    "ONS population estimate" = "#B2182B",
    "Registered patients" = "#2166AC"
  )) +
  scale_x_discrete(labels = c(
    "ONS population estimate" = "ONS population\nestimate",
    "Registered patients" = "Registered\npatients"
  )) +
  scale_y_continuous(
    limits = c(0, NA),
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Patients per qualified GP FTE",
    subtitle = "England, July 2024",
    x = NULL,
    y = "People per qualified GP FTE",
    caption = paste0(
      "Both bars use the same July 2024 qualified GP total: ",
      scales::comma(patients_per_gp_july_2024$GP_FTE[1], accuracy = 0.1),
      " FTE."
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  ),
  "Practice-level registrations",
  "GP FTE: practice-level workforce data; aggregated to the chart geography."
)
```

## July time series

```{r patients-per-gp-july-time-series}
label_chart(
ggplot(
  patients_per_gp_july,
  aes(x = Period, y = Patients_Per_GP_FTE, colour = Source)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population estimate" = "#B2182B"
  )) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(
    limits = c(0, NA),
    labels = scales::label_comma(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Patients per qualified GP FTE",
    subtitle = "England, July snapshots",
    x = NULL,
    y = "Patients per GP FTE",
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ),
  "Practice-level registrations",
  "GP FTE: practice-level workforce data; aggregated to the chart geography."
)
```

```{r patients-per-gp-national-table, results='asis'}
patients_per_gp_national_snapshot %>%
  transmute(
    Geography,
    `Registered patients per GP FTE` =
      scales::comma(Registered_Per_GP_FTE, accuracy = 1),
    `ONS population per GP FTE` =
      scales::comma(ONS_Per_GP_FTE, accuracy = 1)
  ) %>%
  report_table(caption = "National patients per GP FTE comparison (Practice registrations; practice-level workforce)")
```

## NHS England regions

```{r patients-per-gp-region-dumbbell, fig.height=7}
gp_region_dumbbell <- patients_per_gp_region_snapshot %>%
  mutate(
    NHS_Region = reorder(NHS_Region, Registered_Per_GP_FTE)
  )

label_chart(
ggplot(
  gp_region_dumbbell,
  aes(y = NHS_Region)
) +
  geom_segment(
    aes(
      x = ONS_Per_GP_FTE,
      xend = Registered_Per_GP_FTE,
      yend = NHS_Region
    ),
    colour = "grey65",
    linewidth = 0.9
  ) +
  geom_point(
    aes(x = Registered_Per_GP_FTE, colour = "Registered patients"),
    size = 3
  ) +
  geom_point(
    aes(x = ONS_Per_GP_FTE, colour = "ONS population"),
    size = 3
  ) +
  geom_text(
    colour = "grey20",
    aes(
      x = Registered_Per_GP_FTE,
      label = scales::comma(Registered_Per_GP_FTE, accuracy = 1)
    ),
    hjust = -0.12,
    vjust = -0.7,
    size = 3.4
  ) +
  geom_text(
    colour = "grey20",
    aes(
      x = ONS_Per_GP_FTE,
      label = scales::comma(ONS_Per_GP_FTE, accuracy = 1)
    ),
    hjust = -0.12,
    vjust = 1.25,
    size = 3.4
  ) +
  scale_x_continuous(
    labels = scales::label_comma(),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.18))
  ) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population" = "#B2182B"
  )) +
  labs(
    title = "Patients per qualified GP FTE by NHS England region",
    subtitle = format(snapshot_date, "%B %Y"),
    x = "Patients per GP FTE",
    y = NULL,
    colour = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ),
  "Practice-level registrations",
  "GP FTE: practice-level workforce data; aggregated to the chart geography."
)
```

```{r patients-per-gp-region-table, results='asis'}
patients_per_gp_region_snapshot %>%
  arrange(desc(Registered_Per_GP_FTE)) %>%
  transmute(
    `NHS England region` = NHS_Region,
    `Registered patients per GP FTE` =
      scales::comma(Registered_Per_GP_FTE, accuracy = 1),
    `ONS population per GP FTE` =
      scales::comma(ONS_Per_GP_FTE, accuracy = 1)
  ) %>%
  report_table(caption = "Regional patients per GP FTE comparison (Practice registrations; practice-level workforce)")
```

## Integrated care boards

```{r patients-per-gp-icb-dumbbell, fig.height=13}
gp_icb_dumbbell <- patients_per_gp_icb_snapshot %>%
  mutate(
    ICB = reorder(ICB, Registered_Per_GP_FTE)
  )

label_chart(
ggplot(
  gp_icb_dumbbell,
  aes(y = ICB)
) +
  geom_segment(
    aes(
      x = ONS_Per_GP_FTE,
      xend = Registered_Per_GP_FTE,
      yend = ICB
    ),
    colour = "grey65",
    linewidth = 0.8
  ) +
  geom_point(
    aes(x = Registered_Per_GP_FTE, colour = "Registered patients"),
    size = 2.5
  ) +
  geom_point(
    aes(x = ONS_Per_GP_FTE, colour = "ONS population"),
    size = 2.5
  ) +
  geom_text(
    colour = "grey20",
    aes(
      x = Registered_Per_GP_FTE,
      label = scales::comma(Registered_Per_GP_FTE, accuracy = 1)
    ),
    hjust = -0.10,
    vjust = -0.7,
    size = 2.8
  ) +
  geom_text(
    colour = "grey20",
    aes(
      x = ONS_Per_GP_FTE,
      label = scales::comma(ONS_Per_GP_FTE, accuracy = 1)
    ),
    hjust = -0.10,
    vjust = 1.25,
    size = 2.8
  ) +
  scale_x_continuous(
    labels = scales::label_comma(),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.20))
  ) +
  scale_colour_manual(values = c(
    "Registered patients" = "#2166AC",
    "ONS population" = "#B2182B"
  )) +
  labs(
    title = "Patients per qualified GP FTE by NHS integrated care board",
    subtitle = format(snapshot_date, "%B %Y"),
    x = "Patients per GP FTE",
    y = NULL,
    colour = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  ),
  "Practice-level registrations",
  "GP FTE: practice-level workforce data; aggregated to the chart geography."
)
```

```{r patients-per-gp-icb-table, results='asis'}
patients_per_gp_icb_snapshot %>%
  arrange(desc(Registered_Per_GP_FTE)) %>%
  transmute(
    ICB,
    `Registered patients per GP FTE` =
      scales::comma(Registered_Per_GP_FTE, accuracy = 1),
    `ONS population per GP FTE` =
      scales::comma(ONS_Per_GP_FTE, accuracy = 1)
  ) %>%
  report_table(caption = "ICB patients per GP FTE comparison (Practice registrations; practice-level workforce)")
```

```{r payment-scenario-prepare, include=FALSE, results='asis'}
source("13_Model_Payment_Reduction.R")
payment_metric_name <- "Total NHS Payments to General Practice Minus Deductions"

payment_data <- finance_datasets %>%
  mutate(
    Effective_Snapshot_Date = as.Date(Effective_Snapshot_Date),
    Metric_Value = as.numeric(Metric_Value),
    Practice_Code = as.character(Practice_Code)
  ) %>%
  filter(Metric == payment_metric_name)

if (nrow(payment_data) == 0L ||
    any(is.na(payment_data$Effective_Snapshot_Date))) {
  stop("No valid payment snapshot is available for the selected metric.")
}
payment_snapshot_date <- max(payment_data$Effective_Snapshot_Date)
payment_current <- payment_data %>%
  filter(Effective_Snapshot_Date == payment_snapshot_date)
if (any(!is.finite(payment_current$Metric_Value))) {
  stop("The selected payment snapshot contains missing or non-finite amounts.")
}
payment_by_practice <- payment_current %>%
  group_by(Practice_Code) %>%
  summarise(Total_Payments = sum(Metric_Value), .groups = "drop")

national_payment_scenario <- model_payment_reduction(tibble::tibble(
  Geography = "England",
  Total_Payments = sum(payment_by_practice$Total_Payments),
  Registered = national_snapshot$Registered,
  ONS = national_snapshot$ONS
))

regional_payment_scenario <- payment_by_practice %>%
  inner_join(practice_region_lookup, by = "Practice_Code") %>%
  group_by(NHS_Region) %>%
  summarise(Total_Payments = sum(Total_Payments), .groups = "drop") %>%
  left_join(practice_region_registered, by = "NHS_Region") %>%
  left_join(
    region_snapshot %>%
      transmute(NHS_Region = normalise_region_name(NHSER26NM), ONS),
    by = "NHS_Region"
  ) %>%
  model_payment_reduction()

icb_payment_scenario <- payment_by_practice %>%
  inner_join(practice_icb_lookup, by = "Practice_Code") %>%
  group_by(ICB_Key, ICB) %>%
  summarise(Total_Payments = sum(Total_Payments), .groups = "drop") %>%
  rename(ICB_Payment_Name = ICB) %>%
  left_join(
    practice_icb_registered %>% rename(ICB_Registered_Name = ICB),
    by = "ICB_Key"
  ) %>%
  left_join(icb_population_snapshot, by = "ICB_Key") %>%
  mutate(ICB = coalesce(ICB_Population_Name, ICB_Registered_Name, ICB_Payment_Name)) %>%
  model_payment_reduction()

payment_unassigned_region <- payment_by_practice %>%
  anti_join(practice_region_lookup, by = "Practice_Code") %>%
  summarise(Amount = sum(Total_Payments)) %>% pull(Amount)
payment_unassigned_icb <- payment_by_practice %>%
  anti_join(practice_icb_lookup, by = "Practice_Code") %>%
  summarise(Amount = sum(Total_Payments)) %>% pull(Amount)
registered_unassigned_region <- national_snapshot$Registered -
  sum(practice_region_registered$Registered)
registered_unassigned_icb <- national_snapshot$Registered -
  sum(practice_icb_registered$Registered)
if (registered_unassigned_region < -0.5 || registered_unassigned_icb < -0.5) {
  stop("Geographic registration totals exceed the national practice total.")
}

payment_currency <- function(x, accuracy = 1) {
  scales::dollar(x, prefix = gbp_symbol, accuracy = accuracy)
}
payment_loss_chart <- function(data, geography, title) {
  ggplot(data, aes(
    x = Potential_Reduction,
    y = reorder(.data[[geography]], Potential_Reduction)
  )) +
    geom_col(fill = "#2166AC", width = 0.65) +
    geom_text(
    colour = "grey20",
      aes(label = paste0(
        gbp_symbol, scales::number(Potential_Reduction / 1e6, accuracy = 0.1),
        "m (", sprintf("%.1f%%", Reduction_Pct), ")"
      )),
      hjust = -0.08, size = 3.1
    ) +
    scale_x_continuous(
      labels = scales::label_currency(prefix = gbp_symbol, scale = 1e-6, suffix = "m"),
      limits = c(0, NA), expand = expansion(mult = c(0, 0.25))
    ) +
    labs(
      title = title,
      subtitle = paste("Payment snapshot:", format(payment_snapshot_date, "%d %b %Y"),
                       "| Population:", format(snapshot_date, "%B %Y")),
      x = "Modelled payment loss", y = NULL,
      caption = "Labels show money lost and percentage of current payments."
    ) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
}
payment_loss_per_resident_chart <- function(data, geography, title) {
  chart_data <- data %>% filter(is.finite(Loss_Per_ONS_Resident))
  ggplot(chart_data, aes(x = Loss_Per_ONS_Resident,
    y = reorder(.data[[geography]], Loss_Per_ONS_Resident))) +
    geom_col(fill = "#2166AC", width = 0.65) +
    geom_text(aes(label = payment_currency(Loss_Per_ONS_Resident, accuracy = 0.01)),
              colour = "grey20", hjust = -0.08, size = 3.1) +
    scale_x_continuous(labels = function(x) payment_currency(x, accuracy = 1),
                      limits = c(0, NA), expand = expansion(mult = c(0, 0.2))) +
    coord_cartesian(clip = "off") +
    labs(title = title,
      subtitle = paste("Payment snapshot:", format(payment_snapshot_date, "%d %b %Y"),
                       "| ONS population:", format(snapshot_date, "%B %Y")),
      x = "Modelled annual payment loss per ONS resident", y = NULL,
      caption = "Annual modelled loss divided by ONS residents; zero-resident rates are undefined.") +
    theme_minimal(base_size = 11) +
    theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
          plot.margin = margin(6, 25, 6, 6))
}

payment_loss_table <- function(data, geography, caption) {
  data %>%
    arrange(desc(Potential_Reduction)) %>%
    transmute(
      Geography = .data[[geography]],
      `Current (£)` = scales::comma(Total_Payments, accuracy = 1),
      `Rate (£/patient)` = scales::number(Payment_Per_Registered, accuracy = 0.01),
      `List reduction (n)` = scales::comma(Registrations_Removed),
      `Loss (£)` = scales::comma(Potential_Reduction, accuracy = 1),
      `Remaining (£)` = scales::comma(Remaining_Payments, accuracy = 1)
    ) %>%
    report_table(caption = paste(caption, "(Practice registrations; practice-level payments)"))
}
payment_loss_per_resident_table <- function(data, geography, caption) {
  data %>% arrange(desc(Loss_Per_ONS_Resident)) %>%
    transmute(Geography = .data[[geography]],
      `ONS residents` = scales::comma(ONS),
      `Practice registrations` = scales::comma(Registered),
      `Loss (£/ONS resident)` = ifelse(is.na(Loss_Per_ONS_Resident), "Not defined",
        scales::number(Loss_Per_ONS_Resident, accuracy = 0.01)),
      `Loss (£/registered patient)` = scales::number(Loss_Per_Registered_Patient,
                                                    accuracy = 0.01)) %>%
    report_table(caption = caption)
}
```

```{r fixed-total-payment-prepare, include=FALSE}
national_fixed_payment_rates <- model_fixed_total_payment_rates(national_payment_scenario)
regional_fixed_payment_rates <- model_fixed_total_payment_rates(regional_payment_scenario)
icb_fixed_payment_rates <- model_fixed_total_payment_rates(icb_payment_scenario)

fixed_payment_rate_chart <- function(data, geography, title) {
  chart_data <- data %>%
    mutate(Rate_Geography = .data[[geography]]) %>%
    arrange(Payment_Per_Registered)
  chart_data$Rate_Geography <- factor(
    chart_data$Rate_Geography, levels = unique(chart_data$Rate_Geography)
  )
  ggplot(chart_data, aes(y = Rate_Geography)) +
    geom_segment(
      aes(x = Payment_Per_Registered, xend = Payment_Per_ONS, yend = Rate_Geography),
      colour = "grey65", linewidth = 0.8, na.rm = TRUE
    ) +
    geom_point(aes(x = Payment_Per_Registered, colour = "Registered population"), size = 3) +
    geom_point(aes(x = Payment_Per_ONS, colour = "ONS population"), size = 3, na.rm = TRUE) +
    geom_text(
    colour = "grey20",
      aes(x = Payment_Per_Registered,
          label = payment_currency(Payment_Per_Registered, accuracy = 0.01)),
      nudge_y = 0.22, size = 3, show.legend = FALSE
    ) +
    geom_text(
    colour = "grey20",
      aes(x = Payment_Per_ONS,
          label = payment_currency(Payment_Per_ONS, accuracy = 0.01)),
      nudge_y = -0.22, size = 3, na.rm = TRUE, show.legend = FALSE
    ) +
    scale_colour_manual(values = c(
      "Registered population" = "#2166AC", "ONS population" = "#B2182B"
    )) +
    scale_x_continuous(labels = function(x) payment_currency(x, accuracy = 1),
                       expand = expansion(mult = c(0.15, 0.18))) +
    coord_cartesian(clip = "off") +
    labs(
      title = title,
      subtitle = paste("Payment snapshot:", format(payment_snapshot_date, "%d %b %Y"),
                       "| Population:", format(snapshot_date, "%B %Y")),
      x = "Average annual payment per person", y = NULL, colour = "Population denominator",
      caption = "Each geography's total payments stay unchanged; only the denominator changes."
    ) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "top", panel.grid.major.y = element_blank(),
          panel.grid.minor = element_blank(), plot.margin = margin(8, 25, 8, 8))
}

fixed_payment_rate_table <- function(data, geography, caption) {
  data %>%
    arrange(desc(Payment_Per_Registered)) %>%
    transmute(
      Geography = .data[[geography]],
      `Total payments (£)` = scales::comma(Total_Payments, accuracy = 1),
      `Registered, n` = scales::comma(Registered),
      `ONS, n` = scales::comma(ONS),
      `£/registered patient` = scales::number(Payment_Per_Registered, accuracy = 0.01),
      `£/ONS resident` = if_else(is.na(Payment_Per_ONS), "Unavailable",
                                scales::number(Payment_Per_ONS, accuracy = 0.01)),
      `Change (£/person)` = if_else(is.na(Rate_Change), "Unavailable",
                                   scales::number(Rate_Change, accuracy = 0.01)),
      `Change (%)` = if_else(is.na(Rate_Change_Pct), "Not defined",
                             sprintf("%.2f%%", Rate_Change_Pct))
    ) %>%
    report_table(caption = paste(caption, "(Practice registrations; practice-level payments)"))
}
```

# If payment to general practice were to remain constant despite reduction in list size

This comparison keeps overall payments the same but changes the population denominator.

For each geography, the full annual payment total is divided separately by the practice-level registered count and the ONS population. Payments use **Total NHS Payments to General Practice Minus Deductions**, with the latest available payment snapshot (`r format(payment_snapshot_date, "%d %B %Y")`); populations refer to `r format(snapshot_date, "%B %Y")`. Registrations and payments follow practice assignment, while the ONS denominator follows residence; cross-boundary registration can affect local comparisons. ICBs use April 2026 geography. Amounts without a region assignment are `r payment_currency(payment_unassigned_region)`; amounts without an ICB assignment are `r payment_currency(payment_unassigned_icb)` and are excluded from the respective local totals. A higher payment per resident is an average under a fixed budget, not extra income.

```{r national-fixed-total-payment-chart, fig.width=8, fig.height=5}
label_chart(
national_fixed_payment_rates %>%
  select(Payment_Per_Registered, Payment_Per_ONS) %>%
  pivot_longer(everything(), names_to = "Denominator", values_to = "Average_Payment") %>%
  mutate(Denominator = factor(
    Denominator, levels = c("Payment_Per_Registered", "Payment_Per_ONS"),
    labels = c("Registered population", "ONS population")
  )) %>%
  ggplot(aes(x = Denominator, y = Average_Payment, fill = Denominator)) +
  geom_col(width = 0.6, na.rm = TRUE) +
  geom_text(aes(label = payment_currency(Average_Payment, accuracy = 0.01)),
            colour = "grey20",
            vjust = -0.5, size = 4, na.rm = TRUE) +
  scale_fill_manual(values = c(
    "Registered population" = "#2166AC", "ONS population" = "#B2182B"
  )) +
  scale_y_continuous(labels = function(x) payment_currency(x, accuracy = 1),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "National average payment with total payments unchanged",
    subtitle = paste("Payment snapshot:", format(payment_snapshot_date, "%d %b %Y"),
                     "| Population:", format(snapshot_date, "%B %Y")),
    x = "Population denominator", y = "Average annual payment per person",
    caption = "The full national payment total is used in both bars."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", panel.grid.minor = element_blank()),
  "Practice-level registrations",
  "Payments: practice-level annual payment data; aggregated to the chart geography."
)
```

```{r national-fixed-total-payment-table, results='asis'}
fixed_payment_rate_table(national_fixed_payment_rates, "Geography",
                         "National average payments with the total held constant")
```

## Average payment by NHS England region

```{r regional-fixed-total-payment-chart, fig.height=7}
label_chart(
fixed_payment_rate_chart(regional_fixed_payment_rates, "NHS_Region",
                         "Regional average payments with total payments unchanged"),
  "Practice-level registrations",
  "Payments: practice-level annual payment data; aggregated to the chart geography."
)
```

```{r regional-fixed-total-payment-table, results='asis'}
fixed_payment_rate_table(regional_fixed_payment_rates, "NHS_Region",
                         "Regional average payments with each total held constant")
```

## Average payment by integrated care board

```{r icb-fixed-total-payment-chart, fig.height=13}
label_chart(
fixed_payment_rate_chart(icb_fixed_payment_rates, "ICB",
                         "ICB average payments with totals unchanged (April 2026 geography)"),
  "Practice-level registrations",
  "Payments: practice-level annual payment data; aggregated to the chart geography."
)
```

```{r icb-fixed-total-payment-table, results='asis'}
fixed_payment_rate_table(icb_fixed_payment_rates, "ICB",
                         "ICB average payments with each total held constant")
```

# If payments decreased alongside reduction in list size

This section examines how much money could be removed from practices if registered lists were reduced to be in line with the ONS. It uses the calculated payment per capita before removal and multiplies it by the removed number of patients, using Total NHS Payments to General Practice Minus Deductions.

For each geography, the rate is total annual payments divided by practice-level registrations; the list reduction is the positive difference between those registrations and ONS residents. The loss is the rate multiplied by that reduction, and remaining payments are current payments minus the loss. For example, £165 per patient multiplied by 1,000 removed registrations gives a £165,000 loss. Where ONS residents equal or exceed registrations, the loss is zero. Each geography uses its own rate, so local losses need not sum to the national estimate. This is an illustrative proportional-payment scenario, not a prediction of contractual losses. The additional regional and ICB charts divide each geography's modelled annual loss by its July 2024 ONS resident population. This expresses loss per person and adjusts for population size; it does not measure density per square kilometre. The tables show both population denominators and loss per ONS resident and per registered patient. Zero ONS populations produce an undefined per-resident rate rather than zero.

```{r national-payment-loss-chart, fig.height=3.5}
label_chart(
payment_loss_chart(national_payment_scenario, "Geography", "Potential national payment loss"),
  "Practice-level registrations",
  "Payments: practice-level annual payment data; aggregated to the chart geography."
)
```

```{r national-payment-loss-table, results='asis'}
payment_loss_table(national_payment_scenario, "Geography", "National payment loss scenario")
```

## Loss by NHS England region

```{r regional-payment-loss-chart, fig.height=7}
label_chart(
payment_loss_chart(regional_payment_scenario, "NHS_Region",
                   "Potential payment loss by NHS England region"),
  "Practice-level registrations",
  "Payments: practice-level annual payment data; aggregated to the chart geography."
)
```

```{r regional-payment-loss-per-resident-chart, fig.height=7}
label_chart(
  payment_loss_per_resident_chart(regional_payment_scenario, "NHS_Region",
    "Potential annual payment loss per resident by NHS England region"),
  "Practice-level registrations",
  "Payments: practice-level annual payment data. Per-person denominator: ONS residents."
)
```

```{r regional-payment-loss-per-resident-table, results='asis'}
payment_loss_per_resident_table(regional_payment_scenario, "NHS_Region",
  "Population denominators and annual modelled loss per person")
```

```{r regional-payment-loss-table, results='asis'}
payment_loss_table(regional_payment_scenario, "NHS_Region", "Regional payment loss scenarios")
```

## Loss by integrated care board

```{r icb-payment-loss-chart, fig.height=13}
label_chart(
payment_loss_chart(icb_payment_scenario, "ICB",
                   "Potential payment loss by ICB (April 2026 geography)"),
  "Practice-level registrations",
  "Payments: practice-level annual payment data; aggregated to the chart geography."
)
```

```{r icb-payment-loss-per-resident-chart, fig.height=13}
label_chart(
  payment_loss_per_resident_chart(icb_payment_scenario, "ICB",
    "Potential annual payment loss per resident by integrated care board"),
  "Practice-level registrations",
  "Payments: practice-level annual payment data. Per-person denominator: ONS residents."
)
```

```{r icb-payment-loss-per-resident-table, results='asis'}
payment_loss_per_resident_table(icb_payment_scenario, "ICB",
  "Population denominators and annual modelled loss per person")
```

```{r icb-payment-loss-table, results='asis'}
payment_loss_table(icb_payment_scenario, "ICB", "ICB payment loss scenarios")
```
