---
title: "05_Patient_List_Responsive"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

# Patient List Analysis Responsive Piece

This report explores the impact of recent attempts to reduce ghost patients in the GP patients registered dataset

## Introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

knitr::opts_chunk$set(
  fig.width = 15,
  fig.height = 5
)
```

```{r, echo = FALSE}

library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(knitr)
library(stringr)

source("01_connect_to_facts_and_dimensions.R")
source("03_Payments_To_GP_Practice.R")
source("06_Registered_Patients.R")
source("07_Get_LSOA_Data.R")

payments2425 <- read.csv("payments2425.csv", stringsAsFactors = FALSE)

payments_lookup <- payments2425 %>%
  select(
    Practice.Code,
    Region = NHS.England..Region..Name,
    ICB_Name = Sub.ICB.Name
  ) %>%
  distinct(Practice.Code, .keep_all = TRUE)

practice_change <- monthly_practice %>%
  filter(
    Period %in% as.Date(c("2025-10-01", "2026-06-01"))
  ) %>%
  select(GP_Practice_Code, Period, Total_Size) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Size
  ) %>%
  mutate(
    Size_Change = `2026-06-01` - `2025-10-01`
  ) %>%
  arrange(Size_Change)

# ggplot(
#   practice_change,
#   aes(
#     x = reorder(GP_Practice_Code, Size_Change),
#     y = Size_Change
#   )
# ) +
#   geom_col(fill = "steelblue") +
#   geom_hline(yintercept = 0, linetype = 2) +
#   coord_flip() +
#   labs(
#     title = "Change in GP practice list size (trimmed)",
#     subtitle = "Excluding top and bottom 5% of practices",
#     x = "GP Practice",
#     y = "Change in list size"
#   ) +
#   theme_minimal()

```

When looking at the most recent ONS population estimates for LSOA (2024) against the same month Registered Patients (July 2024) we can see that the Registered Patients even then was larger than the ONS estimates.

```{r, echo = FALSE}

LSOA_ONS_renamed <- LSOA_ONS %>%
  rename(
    LSOA_Code = Area_Code,
    ONS_Population = Total_Size
  )%>%
  filter(!grepl("^W0", LSOA_Code))

LSOA_Patient_Reg_2024 <- LSOA_Patient_Reg_2024 %>%
  rename(
    Registered_Patients = Total_Size
  )

LSOA_Comparison <- LSOA_Patient_Reg_2024 %>%
  left_join(
    LSOA_ONS_renamed,
    by = c("LSOA_Code", "Effective_Snapshot_Date")
  )%>%
  filter(!grepl("^W0", LSOA_Code)) %>%
  filter(!grepl("OTHER", LSOA_Code))


LSOA_Comparison <- LSOA_Comparison %>%
  mutate(
    Difference = Registered_Patients - ONS_Population,
    Pct_ONS_of_Registered = 100 * Registered_Patients  / ONS_Population
  )

overall_totals <- LSOA_Comparison %>%
  summarise(
    Total_Registered = sum(Registered_Patients, na.rm = TRUE),
    Total_ONS = sum(ONS_Population, na.rm = TRUE),
    Total_Difference = Total_Registered - Total_ONS,
    Pct_Difference = 100 * Total_Difference / Total_Registered
  ) %>%
  mutate(
    Total_Registered = scales::comma(Total_Registered),
    Total_ONS = scales::comma(Total_ONS),
    Total_Difference = scales::comma(Total_Difference),
    Pct_Difference = paste0(
      round(Pct_Difference, 1), "%"
    )
  )

overall_totals



```

This gap is more noticeable for men with a larger gap between the two datasets compared to women. The ONS estimates there should be more women then men but this is the opposite in registered patients data. Due to the recent data cleaning numbers of men have been reducing faster than women, and in the most recent data they are nearly at the same level, but both remain higher than what the ONS estimates.

```{r, echo = FALSE}

LSOA_ONS_renamed_sex <- LSOA_ONS_SEX %>%
  rename(
    LSOA_Code = Area_Code,
    ONS_Population = Total_Size   
  ) %>%
  filter(!grepl("^W0", LSOA_Code))


LSOA_Patient_Reg_2024_Sex <- LSOA_Patient_Reg_2024_Sex %>%
  rename(
    Registered_Patients = Total_Size 
  ) %>%
  mutate(
    Sex = recode(
      Sex,
      "FEMALE" = "Female",
      "MALE" = "Male"
    )
  )

LSOA_Comparison_Sex <- LSOA_Patient_Reg_2024_Sex %>%
  left_join(
    LSOA_ONS_renamed_sex,
    by = c("LSOA_Code", "Effective_Snapshot_Date", "Sex")
  ) %>%
  filter(
    !grepl("^W0", LSOA_Code),
    !grepl("OTHER", LSOA_Code)
  ) %>%
  mutate(
    Difference = Registered_Patients - ONS_Population,
    Pct_ONS_of_Registered = 100 * Registered_Patients / ONS_Population
  )


# ---- Totals by sex ----
overall_totals_Sex <- LSOA_Comparison_Sex %>%
  group_by(Sex) %>%
  summarise(
    Total_Registered = sum(Registered_Patients, na.rm = TRUE),
    Total_ONS = sum(ONS_Population, na.rm = TRUE),
    Total_Difference = Total_Registered - Total_ONS,
    Pct_Difference = 100 * Total_Difference /  Total_Registered,
    .groups = "drop"
  ) %>%
  mutate(
    Total_Registered = scales::comma(Total_Registered),
    Total_ONS = scales::comma(Total_ONS),
    Total_Difference = scales::comma(Total_Difference),
    Pct_Difference = paste0(
      round(Pct_Difference, 1), "%"
    )
  )

overall_totals_Sex

```

```{r, echo = FALSE}
# Rename ONS columns
LSOA_ONS_Age_renamed <- LSOA_ONS_Age %>%
  rename(
    LSOA_Code = Area_Code,
    ONS_Population = Total_Size
  ) %>%
  filter(!grepl("^W0", LSOA_Code))

# Rename patient register columns
LSOA_Patient_Reg_age_2024 <- LSOA_Patient_Reg_age_2024 %>%
  rename(
    Registered_Patients = Total_Size
  )

# Join the datasets
LSOA_Comparison_Age <- LSOA_Patient_Reg_age_2024 %>%
  left_join(
    LSOA_ONS_Age_group,
    by = c( "Effective_Snapshot_Date", "Age_Band")
  ) %>%
  mutate(
    Difference = Registered_Patients - Total_Size,
    Pct_Difference = 100 * Difference / Registered_Patients
  )

overall_totals_Age <- LSOA_Comparison_Age %>%
  group_by(Age_Band) %>%
  summarise(
    Total_Registered = sum(Registered_Patients, na.rm = TRUE),
    Total_ONS = sum(Total_Size, na.rm = TRUE),
    Total_Difference = Total_Registered - Total_ONS,
    Pct_Difference = 100 * Total_Difference / Total_Registered,
    .groups = "drop"
  ) %>%
  mutate(
    Age_Band = factor(
      Age_Band,
      levels = c(
        "0-10", "11-20", "21-30", "31-40",
        "41-50", "51-60", "61-70", "71-80", "81+"
      )
    )
  ) %>%
  arrange(Age_Band) %>%
  mutate(
    Total_Registered = comma(Total_Registered),
    Total_ONS = comma(Total_ONS),
    Total_Difference = comma(Total_Difference),
    Pct_Difference = paste0(round(Pct_Difference, 1), "%")
  )

overall_totals_Age

overall_totals_Age %>%
  mutate(
    Pct_Difference = parse_number(Pct_Difference)
  ) %>%
  ggplot(aes(x = Age_Band, y = Pct_Difference)) +
  geom_col(fill = "#2C7FB8") +
  labs(
    x = "Age band",
    y = "% Difference",
    title = "Percentage difference between registered patients and ONS population by age band"
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
```{r, echo = FALSE}
# Rename ONS columns
LSOA_ONS_Age_Sex_renamed <- LSOA_ONS_Age_Sex_group %>%
  rename(
    ONS_Population = Total_Size
  )

# Rename patient register columns
LSOA_Patient_Reg_age_sex_2024 <- LSOA_Patient_Reg_age_sex_2024 %>%
  rename(
    Registered_Patients = Total_Size
  )

# Join datasets
LSOA_Comparison_Age_Sex <- LSOA_Patient_Reg_age_sex_2024 %>%
  left_join(
    LSOA_ONS_Age_Sex_renamed,
    by = c("Effective_Snapshot_Date", "Sex", "Age_Band")
  ) %>%
  mutate(
    Difference = Registered_Patients - ONS_Population,
    Pct_Difference = 100 * Difference / Registered_Patients
  )

# Totals by age band and sex
overall_totals_Age_Sex <- LSOA_Comparison_Age_Sex %>%
  group_by(Sex, Age_Band) %>%
  summarise(
    Total_Registered = sum(Registered_Patients, na.rm = TRUE),
    Total_ONS = sum(ONS_Population, na.rm = TRUE),
    Total_Difference = Total_Registered - Total_ONS,
    Pct_Difference = 100 * Total_Difference / Total_Registered,
    .groups = "drop"
  ) %>%
  mutate(
    Age_Band = factor(
      Age_Band,
      levels = c(
        "0-10", "11-20", "21-30", "31-40",
        "41-50", "51-60", "61-70", "71-80", "81+"
      )
    )
  ) %>%
  arrange(Age_Band, Sex ) %>%
  mutate(
    Total_Registered = comma(Total_Registered),
    Total_ONS = comma(Total_ONS),
    Total_Difference = comma(Total_Difference),
    Pct_Difference = paste0(round(Pct_Difference, 1), "%")
  )

overall_totals_Age_Sex

overall_totals_Age_Sex %>%
  mutate(
    Pct_Difference = parse_number(Pct_Difference)
  ) %>%
  ggplot(aes(x = Age_Band, y = Pct_Difference, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c("Female" = "#E78AC3", "Male" = "#4DAF4A")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x = "Age band",
    y = "% Difference",
    fill = "Sex",
    title = "Percentage difference between registered patients and ONS population by age band and sex"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

This map shows percentage difference between what the ONS estimates the population is in a LSOA and what the registered patients are as of July 2024. (Showing proportion of registered patients compared to ONS estimates)

```{r, echo = FALSE}
library(sf)
LSOA_shp <- st_read("LSOA_2021_EW_BSC_V4.shp")

LSOA_map <- LSOA_shp %>%
  left_join(
    LSOA_Comparison,
    by = c("LSOA21CD" = "LSOA_Code")
  ) %>%
  filter(!grepl("^W0", LSOA21CD)) 

LSOA_map <- LSOA_map %>%
  mutate(
    Pct_band = case_when(
      Pct_ONS_of_Registered <= 75 ~ "0–75",
      Pct_ONS_of_Registered <= 150 ~ "75–150",
      Pct_ONS_of_Registered <= 250 ~ "150–250",
      Pct_ONS_of_Registered <= 500 ~ "251–500",
      Pct_ONS_of_Registered <= 5000 ~ "501-5000",
      TRUE ~ NA_character_
    )
  )%>%
  filter(!is.na(Pct_band))

  
 ggplot(LSOA_map) +
  geom_sf(
    aes(fill = Pct_band),
    colour = "grey80",
    linewidth = 0.0001
  ) +
  scale_fill_manual(
    values = c(
      "0–75" = "black",
      "75–150" = "white",
      "150–250" = "orange",
      "251–500" = "red",
      "501-5000" = "blue"
    ),
    name = "% ONS of Registered"
  ) +
  theme_void()
```

Looking at individual practices across October 25 to June 26 we can see there has been some variation. With the top 10 and bottom 10 practices by % change highlighted below. Some of this will be caused by mergers which requires further exploration.

```{r, echo = FALSE}

finances_unique <- finance_datasets %>%
  filter(Metric == "Average payments per registered patient") %>%
  distinct(Practice_Code, .keep_all = TRUE)

monthly_practice_finance <- monthly_practice %>%
  left_join(
    finances_unique,
    by = c("GP_Practice_Code" = "Practice_Code")
  )

monthly_practice_finance <- monthly_practice_finance %>%
  left_join(
    payments_lookup,
    by = c("GP_Practice_Code" = "Practice.Code")
  )

monthly_practice_finance <- monthly_practice_finance %>%
  mutate(
    ICB_Name = str_replace(ICB_Name, "ICB.*$", "ICB")
  )

gp_change <- monthly_practice_finance %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  group_by(GP_Practice_Code, Region, Period) %>%
  summarise(
    Total_Size = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Period = format(Period, "%b %Y")
  ) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Size
  ) %>%
  mutate(
    `Oct 2025` = as.numeric(`Oct 2025`),
    `Jun 2026` = as.numeric(`Jun 2026`),
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`
  )

gp_change_top_bottom <- gp_change %>%
  filter(!is.na(`Number change`), is.finite(`Percent change`))

top_10 <- gp_change_top_bottom %>%
  arrange(desc(`Number change`)) %>%
  slice_head(n = 10) %>%
  mutate(Group = "Top 10 increases")

bottom_10 <- gp_change_top_bottom %>%
  arrange(`Number change`) %>%
  slice_head(n = 10) %>%
  mutate(Group = "Bottom 10 decreases")

gp_top_bottom_table <- bind_rows(top_10, bottom_10) %>%
  select(
    Group,
    Region,
    GP_Practice_Code,
    `Oct 2025`,
    `Jun 2026`,
    `Number change`,
    `Percent change`
  ) %>%
  mutate(
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percent change` = scales::percent(`Percent change`, accuracy = 0.1)
  )

knitr::kable(
  gp_top_bottom_table,
  caption = "Top 10 increases and bottom 10 decreases in GP practice list size (Oct 2025 to Jun 2026)",
  align = "l"
)

```

Total GP List size has seen an increase over time until it's peak in November 2025 at 63,947,100, after this total list size has seen a continued decrease and currently stands at 63,488,912 as of June 2026.

```{r, echo = FALSE}
ggplot(
  monthly_overall,
  aes(
    x = Period,
    y = Total_Size
  )
) +
  geom_line(linewidth = 1.2, colour = "steelblue") +
  geom_point(size = 2, colour = "steelblue") +
  scale_x_date(
    date_breaks = "2 months",
    date_labels = "%b\n%Y"
  ) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "m")
  ) +
  labs(
    title = "Total GP list size",
    subtitle = "June 2023 to June 2026",
    x = NULL,
    y = "Total list size"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


# # Month-on-month change table
# monthly_overall %>%
#   arrange(Period) %>%
#   mutate(
#     `Number change` = Total_Size - lag(Total_Size),
#     `% change` = (Total_Size - lag(Total_Size)) / lag(Total_Size)
#   ) %>%
#   transmute(
#     Month = format(Period, "%b %Y"),
#     `Total list size` = comma(Total_Size),
#     `Number change` = comma(`Number change`),
#     `% change` = percent(`% change`, accuracy = 0.01)
#   ) %>%
#   kable(
#     align = c("l", "r", "r", "r"),
#     caption = "Month-on-month change in GP registered population"
#   )

```

Focusing on the peak at November 2025 we can see that the list size has dropped 458,188 patients as of June 2026 a 0.72% decrease.

```{r, echo = FALSE}

overall_from_practices <- monthly_practice %>%
  filter(
    Period >= as.Date("2025-10-01"),
    Period <= as.Date("2026-06-30")
  ) %>%
  group_by(Period) %>%
  summarise(
    Total_Patients = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Period)

ggplot(overall_from_practices, aes(x = Period, y = Total_Patients)) +
  geom_line(linewidth = 1.3, colour = "steelblue") +
  geom_point(size = 2.5, colour = "steelblue") +
  scale_y_continuous(
    breaks = 25.2e6,
    labels = scales::label_number(scale = 1e-6, suffix = "m")
  ) +
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b %Y"
  ) +
  labs(
    title = "Total GP registered population",
    subtitle = "October 2025 to June 2026 (summed from practice-level data)",
    x = NULL,
    y = "Total registered patients"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

Trying to match similar work done by Helen Salisbury in the BMJ looking at the change between November 2025 and May 2026 we can match the 344,000 patients removed they were referencing, and a similar cost if using a global sum payment of 121.79.

```{r, echo = FALSE}
overall_change_table <- overall_from_practices %>%
  filter(Period %in% as.Date(c("2025-11-01", "2026-05-01"))) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  select(Period, Total_Patients) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    Change = `May 2026` - `Nov 2025`,
    Cost = Change * 121.79,
    `% change` = Change / `Nov 2025`
  )

overall_change_table <- overall_change_table %>%
  mutate(
    `Nov 2025` = comma(`Nov 2025`),
    `May 2026` = comma(`May 2026`),
    Change = comma(Change),
    `% change` = percent(`% change`, accuracy = 0.01),
    Cost = label_currency(prefix = "£", accuracy = 1)(Cost)
  )

overall_change_table <- overall_change_table %>%
  select(`Nov 2025`, `May 2026`, Change, `% change`, Cost)

overall_change_table[] <- lapply(overall_change_table, function(x) {
  if (is.character(x)) {
    iconv(x, from = "", to = "UTF-8", sub = "")
  } else {
    x
  }
})

kable(
  overall_change_table,
  align = c("r", "r", "r", "r", "r"),
  caption = "Overall change in GP registered population and cost (Dec 2025 to May 2026)"
)
```

When grouping practices by their list size in October 2025 we can see that smaller practices have seen a 1.5% decrease over the time period while practices with 20k+ patients have seen an increase of 0.4% in this time period.

```{r, echo = FALSE}


practice_bands <- monthly_practice_finance %>%
  filter(Period == as.Date("2025-10-01")) %>%
  select(GP_Practice_Code, Total_Size) %>%
  mutate(
    Size_Band = case_when(
      Total_Size <= 5000  ~ "0-5,000",
      Total_Size <= 10000 ~ "5,001-10,000",
      Total_Size <= 15000 ~ "10,001-15,000",
      Total_Size <= 20000 ~ "15,001-20,000",
      TRUE                ~ "20,000+"
    )
  ) %>%
  select(GP_Practice_Code, Size_Band)

monthly_bands <- monthly_practice_finance %>%
  left_join(practice_bands, by = "GP_Practice_Code") %>%
  filter(!is.na(Size_Band))

band_trends <- monthly_bands %>%
  group_by(Size_Band, Period) %>%
  summarise(
    Sum_Practice_Size = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  )

# Create summary table
band_change <- band_trends %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  select(Size_Band, Period, Sum_Practice_Size) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Sum_Practice_Size
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`,
    Size_Band = factor(
      Size_Band,
      levels = c(
        "0-5,000",
        "5,001-10,000",
        "10,001-15,000",
        "15,001-20,000",
        "20,000+"
      )
    )
  ) %>%
  arrange(Size_Band)

# Horizontal bar chart of percentage change
ggplot(
  band_change,
  aes(
    x = `Percent change`,
    y = Size_Band,
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.1)),
    hjust = -0.15,
    size = 4
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.15))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  labs(
    title = "Percentage change in registered patients by October 2025 practice size band",
    subtitle = "October 2025 to June 2026",
    x = "Percentage change",
    y = "October 2025 practice size band"
  ) +
  theme_minimal(base_size = 12)

# Formatted table
band_change %>%
  transmute(
    `October 2025 size band` = Size_Band,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.1)
  ) %>%
  kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by October 2025 practice size band (October 2025 to June 2026)"
  )
```

When looking at list size by practice contract type we can see that practices with APMS contracts have seen the largest decrease losing 82,742 patients or 3.1% of their list size.

```{r, echo = FALSE}

contract_trends <- monthly_practice_finance %>%
    filter(!is.na(Contract_Type)) %>%
  group_by(Contract_Type, Period) %>%
  summarise(
    Total_Patients = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  )

# Create summary table
contract_change <- contract_trends %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  select(Contract_Type, Period, Total_Patients) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`
  ) %>%
  arrange(desc(`Percent change`))

# Horizontal bar chart of percentage change
ggplot(
  contract_change,
  aes(
    x = `Percent change`,
    y = reorder(Contract_Type, `Percent change`),
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.1)),
    hjust = ifelse(contract_change$`Percent change` >= 0, -0.15, 1.15),
    size = 4
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.15))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  labs(
    title = "Percentage change in registered patients by contract type",
    subtitle = "October 2025 to June 2026",
    x = "Percentage change",
    y = "Contract type"
  ) +
  theme_minimal(base_size = 12)

# Formatted table
contract_change %>%
  transmute(
    `Contract type` = Contract_Type,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.1)
  ) %>%
  kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by contract type (October 2025 to June 2026)"
  )

```

List size for non-dispensing practices saw the majority of the list size change. Losing 426,471 or 0.8% of their list size. While dispensing practices lost 0.02%. This means out of all the patients removed non-dispensing practices have seen 99.5% the removals.

```{r, echo = FALSE}
Dispensing_trends <- monthly_practice_finance %>%
    filter(!is.na(Dispensing_Practice)) %>%
  group_by(Dispensing_Practice, Period) %>%
  summarise(
    Total_Patients = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  )

# Create summary table
dispensing_change <- Dispensing_trends %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  select(Dispensing_Practice, Period, Total_Patients) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`
  ) %>%
  arrange(desc(`Percent change`))

# Horizontal bar chart of percentage change
ggplot(
  dispensing_change,
  aes(
    x = `Percent change`,
    y = reorder(Dispensing_Practice, `Percent change`),
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.1)),
    hjust = ifelse(dispensing_change$`Percent change` >= 0, -0.15, 1.15),
    size = 4
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.15))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  labs(
    title = "Percentage change in registered patients by dispensing status",
    subtitle = "October 2025 to June 2026",
    x = "Percentage change",
    y = "Dispensing status"
  ) +
  theme_minimal(base_size = 12)

# Formatted table
dispensing_change %>%
  transmute(
    `Dispensing status` = Dispensing_Practice,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.1)
  ) %>%
  kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by dispensing status (October 2025 to June 2026)"
  )
```

List size for urban practices also saw the majority of the list size change. Losing 428,846 or 0.8% of their list size.

```{r, echo = FALSE}
Rurality_trends <- monthly_practice_finance %>%
      filter(!is.na(Practice_Rurality)) %>%
  group_by(Practice_Rurality, Period) %>%
  summarise(
    Total_Patients = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  )

# Create summary table
rurality_change <- Rurality_trends %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  select(Practice_Rurality, Period, Total_Patients) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`
  ) %>%
  arrange(desc(`Percent change`))

# Horizontal bar chart of percentage change
ggplot(
  rurality_change,
  aes(
    x = `Percent change`,
    y = reorder(Practice_Rurality, `Percent change`),
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.1)),
    hjust = ifelse(rurality_change$`Percent change` >= 0, -0.15, 1.15),
    size = 4
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.15))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  labs(
    title = "Percentage change in registered patients by practice rurality",
    subtitle = "October 2025 to June 2026",
    x = "Percentage change",
    y = "Practice rurality"
  ) +
  theme_minimal(base_size = 12)

# Formatted table
rurality_change %>%
  transmute(
    `Practice rurality` = Practice_Rurality,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.1)
  ) %>%
  kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by practice rurality (October 2025 to June 2026)"
  )

```

Using GP IMD deciles from fingertips which uses Patients Registered at a GP practice LSOA level data to map patient LSOA IMD to create a practice level. We can see that in general more deprived practices have been losing a higher % of their list size over this time period.

```{r, echo = FALSE}

GP_IMD_deciles <- read.csv("GP_IMD_deciles.csv", stringsAsFactors = FALSE)

GP_IMD_deciles <- GP_IMD_deciles %>%
  mutate(
    IMD_Decile = 11 - ntile(Value, 10)
  )

monthly_practice_finance_IMD <- monthly_practice_finance %>%
  left_join(
    GP_IMD_deciles,
    by = c("GP_Practice_Code" = "AreaCode")
  )

IMD_trends <- monthly_practice_finance_IMD %>%
  filter(!is.na(IMD_Decile)) %>%
  group_by(IMD_Decile, Period) %>%
  summarise(
    Total_Patients = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  )


# Create summary table
IMD_change <- IMD_trends %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  select(IMD_Decile, Period, Total_Patients) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`,
    IMD_Decile = factor(
      IMD_Decile,
      levels = 10:1,
      labels = c(
        "10 (Least deprived)",
        "9",
        "8",
        "7",
        "6",
        "5",
        "4",
        "3",
        "2",
        "1 (Most deprived)"
      )
    )
  )

# Horizontal bar chart of percentage change
ggplot(
  IMD_change,
  aes(
    x = `Percent change`,
    y = IMD_Decile,
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.1)),
    hjust = ifelse(IMD_change$`Percent change` >= 0, -0.15, 1.15),
    size = 4
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.02, 0.15))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  labs(
    title = "Percentage change in registered patients by IMD decile",
    subtitle = "October 2025 to June 2026",
    x = "Percentage change",
    y = "IMD decile"
  ) +
  theme_minimal(base_size = 12)

# Formatted table
IMD_change %>%
  transmute(
    `IMD decile` = IMD_Decile,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.1)
  ) %>%
  knitr::kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by IMD decile (October 2025 to June 2026)"
  )


```

The London region has lost a larger percentage of it's list size (1.5%) in this time period compared to other regions.

```{r, echo = FALSE}

Region_trends <- monthly_practice_finance %>%
  filter(!is.na(Region)) %>%
  group_by(Region, Period) %>%
  summarise(
    Total_Patients = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  )


# Create summary table
region_change <- Region_trends %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  select(Region, Period, Total_Patients) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`
  ) %>%
  arrange(desc(`Percent change`))

# Horizontal bar chart of percentage change
ggplot(
  region_change,
  aes(
    x = `Percent change`,
    y = reorder(Region, `Percent change`),
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.1)),
    hjust = ifelse(region_change$`Percent change` >= 0, -0.15, 1.15),
    size = 4
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.05, 0.15))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Percentage change in registered patients by NHS region",
    subtitle = "October 2025 to June 2026",
    x = "Percentage change",
    y = "NHS region"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.margin = margin(5.5, 20, 5.5, 25)
  )

# Formatted table (ordered from largest to smallest percentage change)
region_change %>%
  transmute(
    Region,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.1)
  ) %>%
  knitr::kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by NHS region (October 2025 to June 2026)"
  )
```

```{r, echo = FALSE}

ICB_trends <- monthly_practice_finance %>%
  filter(!is.na(ICB_Name)) %>%
  group_by(ICB_Name, Period) %>%
  summarise(
    Total_Patients = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  )
# ---- Calculate change ----
icb_change <- ICB_trends %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  select(ICB_Name, Period, Total_Patients) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`
  )

# ---- Keep top 5 + bottom 5 by % change ----
icb_top_bottom <- icb_change %>%
  arrange(`Percent change`) %>%
  slice(c(1:5, (n() - 4):n())) %>%
  mutate(
    ICB_Name = factor(ICB_Name, levels = ICB_Name)
  )

# ---- Plot ----
ggplot(
  icb_top_bottom,
  aes(
    x = `Percent change`,
    y = reorder(ICB_Name, `Percent change`),
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = percent(`Percent change`, accuracy = 0.1)),
    hjust = ifelse(icb_top_bottom$`Percent change` >= 0, -0.15, 1.15),
    size = 3.8
  ) +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0.05, 0.2))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "ICBs with the largest and smallest % change in registered patients",
    subtitle = "Top 5 and bottom 5 ICBs (Oct 2025 to Jun 2026)",
    x = "Percentage change",
    y = "ICB"
  ) +
  theme_minimal(base_size = 12)

# ---- Table ----
icb_top_bottom %>%
  transmute(
    `ICB` = ICB_Name,
    `Oct 2025` = comma(`Oct 2025`),
    `Jun 2026` = comma(`Jun 2026`),
    `Number change` = comma(`Number change`),
    `Percentage change` = percent(`Percent change`, accuracy = 0.1)
  ) %>%
  kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Top and bottom 5 ICBs by percentage change in registered patients (Oct 2025 to Jun 2026)"
  )
```

```{r, echo = FALSE}
baseline <- monthly_practice_finance %>%
  filter(Metric == "Average payments per registered patient") %>%
  filter(Period == min(Period, na.rm = TRUE)) %>%
  group_by(GP_Practice_Code) %>%
  summarise(
    Payment = mean(Metric_Value, na.rm = TRUE),
    .groups = "drop"
  )

# assign deciles
baseline <- baseline %>%
  mutate(Payment_Decile = ntile(Payment, 10))

# get ranges for legend labels
decile_labels <- baseline %>%
  group_by(Payment_Decile) %>%
  summarise(
    min_pay = min(Payment, na.rm = TRUE),
    max_pay = max(Payment, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Decile_Label = paste0(
      "D", Payment_Decile,
      " (£", round(min_pay, 2),
      "–£", round(max_pay, 2), ")"
    )
  )

monthly_deciles <- monthly_practice_finance %>%
  filter(Metric == "Average payments per registered patient") %>%
  left_join(baseline, by = "GP_Practice_Code") %>%
  left_join(decile_labels, by = "Payment_Decile")

decile_trends <- monthly_deciles %>%
  filter(!is.na(Payment_Decile)) %>%
  group_by(Decile_Label, Period) %>%
  summarise(
    Total_Patients = sum(Total_Size, na.rm = TRUE),
    .groups = "drop"
  )

decile_trends <- decile_trends %>%
  mutate(
    Decile_Label = iconv(Decile_Label, from = "", to = "UTF-8", sub = "")
  )

# ---- Calculate decile change ----
decile_change <- decile_trends %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  select(Decile_Label, Period, Total_Patients) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`,
    Decile = as.numeric(sub("D([0-9]+).*", "\\1", Decile_Label))
  ) %>%
  arrange(Decile)

# ---- Chart: % change ----
ggplot(
  decile_change,
  aes(
    x = `Percent change`,
    y = reorder(Decile_Label, Decile),
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.1)),
    hjust = ifelse(decile_change$`Percent change` >= 0, -0.15, 1.15),
    size = 3.8
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.05, 0.2))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Percentage change in registered patients by payment-per-patient decile",
    subtitle = "October 2025 to June 2026",
    x = "Percentage change",
    y = "Payment decile"
  ) +
  theme_minimal(base_size = 12)

# ---- Table ----
decile_change %>%
  transmute(
    `Payment decile` = Decile_Label,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.1)
  ) %>%
  knitr::kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by payment-per-patient decile (October 2025 to June 2026)"
  )

```

When looking at the total list size by gender we can see that male patients have been the most impacted. Losing 1% of their numbers compared to 0.3% for women in the same time period.

```{r, echo = FALSE}
gender_long <- monthly_overall %>%
  select(Period, Male_Count, Female_Count) %>%
  pivot_longer(
    cols = c(Male_Count, Female_Count),
    names_to = "Gender",
    values_to = "Count"
  )

gender_long_filtered <- gender_long %>%
  filter(Period >= as.Date("2025-10-01"))

ggplot(gender_long_filtered, aes(x = Period, y = Count, colour = Gender)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(
    labels = scales::label_number(scale = 1e-6, suffix = "m")
  ) +
  labs(
    title = "Registered patients by gender over time",
    x = "Period",
    y = "Number of patients",
    colour = "Gender"
  ) +
  theme_minimal()

gender_change <- gender_long %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Count
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `% change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`
  ) %>%
  mutate(
    `Oct 2025` = comma(`Oct 2025`),
    `Jun 2026` = comma(`Jun 2026`),
    `Number change` = comma(`Number change`),
    `% change` = percent(`% change`, accuracy = 0.1)
  )

kable(
  gender_change,
  align = c("l", "r", "r", "r", "r"),
  caption = "Change in registered patients by gender (October 2025 to June 2026)"
)
```

```{r, echo = FALSE}
age_long <- monthly_overall %>%
  select(
    Period,
    Age_0_9, Age_10_19, Age_20_29, Age_30_39,
    Age_40_49, Age_50_59, Age_60_69, Age_70_79,
    Age_80_89, Age_90_plus
  ) %>%
  pivot_longer(
    cols = -Period,
    names_to = "Age_Group",
    values_to = "Count"
  )

age_long$Age_Group <- factor(
  age_long$Age_Group,
  levels = c(
    "Age_0_9", "Age_10_19", "Age_20_29", "Age_30_39",
    "Age_40_49", "Age_50_59", "Age_60_69", "Age_70_79",
    "Age_80_89", "Age_90_plus"
  )
)

age_long <- age_long %>%
  filter(Period >= as.Date("2025-10-01"))

# ---- Calculate age change ----
age_change <- age_long %>%
  filter(Period %in% as.Date(c("2025-10-01", "2026-06-01"))) %>%
  mutate(Period = format(Period, "%b %Y")) %>%
  pivot_wider(
    names_from = Period,
    values_from = Count
  ) %>%
  mutate(
    `Number change` = `Jun 2026` - `Oct 2025`,
    `Percent change` = (`Jun 2026` - `Oct 2025`) / `Oct 2025`
  ) %>%
  mutate(
    Age_Group = factor(
      Age_Group,
      levels = c(
        "Age_0_9", "Age_10_19", "Age_20_29", "Age_30_39",
        "Age_40_49", "Age_50_59", "Age_60_69", "Age_70_79",
        "Age_80_89", "Age_90_plus"
      )
    )
  ) %>%
  arrange(Age_Group)

# ---- Chart: % change (fixed order) ----
ggplot(
  age_change,
  aes(
    x = `Percent change`,
    y = Age_Group,
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.1)),
    hjust = ifelse(age_change$`Percent change` >= 0, -0.15, 1.15),
    size = 3.8
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.05, 0.2))
  ) +
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Percentage change in registered patients by age group",
    subtitle = "October 2025 to June 2026",
    x = "Percentage change",
    y = "Age group"
  ) +
  theme_minimal(base_size = 12)

# ---- Table ----
age_change %>%
  transmute(
    `Age group` = Age_Group,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Jun 2026` = scales::comma(`Jun 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.1)
  ) %>%
  knitr::kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by age group (October 2025 to June 2026)"
  )
```

```{r, echo = FALSE}
LSOA_joined <- LSOA_Patient_Reg %>%
  left_join(LSOA_IMD, by = "LSOA_Code") %>%
  filter(!startsWith(LSOA_Code, "W0"))

IMD_trends <- LSOA_joined %>%
  filter(!is.na(IMD_Decile)) %>%
  group_by(IMD_Decile, Effective_Snapshot_Date.x) %>%
  summarise(
    Total_Patients = sum(Size, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Calculate change ----
imd_change <- IMD_trends %>%
  filter(Effective_Snapshot_Date.x %in% as.Date(c("2025-10-01", "2026-04-01"))) %>%
  mutate(Period = format(Effective_Snapshot_Date.x, "%b %Y")) %>%
  select(IMD_Decile, Period, Total_Patients) %>%
  pivot_wider(
    names_from = Period,
    values_from = Total_Patients
  ) %>%
  mutate(
    `Number change` = `Apr 2026` - `Oct 2025`,
    `Percent change` = (`Apr 2026` - `Oct 2025`) / `Oct 2025`
  ) %>%
  mutate(
    IMD_Decile = factor(
      IMD_Decile,
      levels = 1:10,
      labels = c(
        "1 (Most deprived)",
        "2",
        "3",
        "4",
        "5",
        "6",
        "7",
        "8",
        "9",
        "10 (Least deprived)"
      )
    )
  ) %>%
  arrange(IMD_Decile)

# ---- % change chart ----
ggplot(
  imd_change,
  aes(
    x = `Percent change`,
    y = IMD_Decile,
    fill = `Percent change`
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::percent(`Percent change`, accuracy = 0.01)),
    hjust = ifelse(imd_change$`Percent change` >= 0, -0.15, 1.15),
    size = 3.8
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0.05, 0.2))
  ) +
  scale_y_discrete(limits = rev(levels(imd_change$IMD_Decile)))+
  scale_fill_gradient2(
    low = "#d73027",
    mid = "grey85",
    high = "#1a9850",
    midpoint = 0,
    guide = "none"
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Percentage change in registered patients by IMD decile",
    subtitle = "October 2025 to April 2026",
    x = "Percentage change",
    y = "IMD decile"
  ) +
  theme_minimal(base_size = 12)

# ---- Table ----
imd_change %>%
  transmute(
    `IMD decile` = IMD_Decile,
    `Oct 2025` = scales::comma(`Oct 2025`),
    `Apr 2026` = scales::comma(`Apr 2026`),
    `Number change` = scales::comma(`Number change`),
    `Percentage change` = scales::percent(`Percent change`, accuracy = 0.01)
  ) %>%
  knitr::kable(
    align = c("l", "r", "r", "r", "r"),
    caption = "Change in registered patients by IMD decile (October 2025 to April 2026)"
  )


```

```{r, echo = FALSE}
# 
# monthly_filtered <- monthly_practice %>%
#   filter(
#     Period >= as.Date("2025-10-01"),
#     Period <= as.Date("2026-06-30")
#   ) %>%
#   mutate(
#     GP_Practice_Code = trimws(as.character(GP_Practice_Code)),
#     Period = as.Date(Period)
#   ) %>%
#   arrange(Period)
# 
# monthly_sets <- monthly_filtered %>%
#   group_by(Period) %>%
#   summarise(
#     practices = list(unique(GP_Practice_Code)),
#     .groups = "drop"
#   ) %>%
#   arrange(Period)
# 
# practice_transitions <- monthly_sets %>%
#   mutate(
#     prev_practices = lag(practices),
#     gained_practices = map2(
#       practices,
#       prev_practices,
#       ~ if (is.null(.y)) character(0) else setdiff(.x, .y)
#     ),
#     lost_practices = map2(
#       practices,
#       prev_practices,
#       ~ if (is.null(.y)) character(0) else setdiff(.y, .x)
#     )
#   )
# 
# 
# lost_df <- practice_transitions %>%
#   select(Period, lost_practices) %>%
#   unnest(lost_practices) %>%
#   rename(GP_Practice_Code = lost_practices) %>%
#   mutate(
#     event = "lost",
#     Previous_Period = Period %m-% months(1)
#   )
# 
# lost_practice_patients <- lost_df %>%
#   left_join(
#     monthly_filtered %>%
#       select(
#         Period,
#         GP_Practice_Code,
#         Total_Size
#       ),
#     by = c(
#       "Previous_Period" = "Period",
#       "GP_Practice_Code"
#     )
#   ) %>%
#   rename(
#     Lost_Period = Period
#   ) %>%
#   select(
#     Lost_Period,
#     Previous_Period,
#     GP_Practice_Code,
#     Total_Size
#   ) %>%
#   arrange(Lost_Period, desc(Total_Size))
# ```
# 
# ```{r}
# {r}
# 
# practice_growth <- monthly_filtered %>%
#   filter(Period %in% c(as.Date("2025-10-01"), as.Date("2026-06-01"))) %>%
#   select(Period, GP_Practice_Code, Total_Size) %>%
#   tidyr::pivot_wider(
#     names_from = Period,
#     values_from = Total_Size,
#     names_prefix = "Size_"
#   ) %>%
#   rename(
#     Baseline_Size = `Size_2025-10-01`,
#     Final_Size    = `Size_2026-06-01`
#   ) %>%
#   mutate(
#     Absolute_Change = Final_Size - Baseline_Size,
#     Percent_Change = 100 * (Final_Size - Baseline_Size) / Baseline_Size
#   )
# 
# # Practices with >=10% growth
# practice_growth_10 <- practice_growth %>%
#   filter(
#     !is.na(Baseline_Size),
#     !is.na(Final_Size),
#     Percent_Change >= 10
#   ) %>%
#   arrange(desc(Percent_Change))
```
